---
title: "AFLResearchProject_OnlineCopy"
author: "Yaron Faustin"
date: "2026-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 4.1 / 4.2
## Load in the data + libraries

Pretty self explanatory

```{r library}
library(readxl) #Used for read.csv
library(lubridate) #Used for ms
library(ggplot2) #GGplot
library(MASS) #fitdistr
library(fitdistrplus) #mledist
library(dplyr) # %>%
library(tidyr) #pivot_widers
library(scales)
library(pheatmap) # special heatmaps

library(grid)
library(mixtools) #for mixture fits
library(nnet) # testing markov assumption
library(stringr) # long-run
library(tibble) # most common transitions
```

load in that data from 2021

```{r data}
#Load in 2021 data
df1 <- read.csv("C:/~/AFL Play by play 1-12.csv")
df2 <- read.csv("C:/~/AFL Play by play 13-27.csv")

playbyplay <- rbind(df1, df2)
```

## Build pbp with all necessary features

```{r build pbp}
#TimeSec and dt
pbp <- playbyplay %>%
  mutate(
    TimeSec = as.numeric(ms(Time))
  ) %>%
  group_by(Match_id, Quarter) %>%
  mutate(
    dt = lead(TimeSec) - TimeSec 
  ) %>%
  ungroup()

#new_period
pbp <- pbp %>%
  group_by(Match_id) %>%
  mutate(
    new_period = Quarter != lag(Quarter, default = first(Quarter)),
    new_period = ifelse(row_number() == 1, TRUE, new_period)
  ) %>%
  ungroup()

# pos
neutral_events <- c(
  "Tackle","Spoil","Hitout",
  "Hitout To Advantage","F 50 Tackle"
)

possession_events <- c(
  "Poss Gain","Stoppage Clearance","Centre Clearance",
  "Kick In","Intercept Mark","Kick","Handball",
  "Free For","Mark","Ground Ball Get",
  "Contested Possession","Uncontested Possession"
)

pbp <- pbp %>%
  mutate(
    pos = case_when(
      Stat %in% neutral_events ~ "Neutral",
      Stat %in% possession_events ~ Team_status,
      TRUE ~ NA_character_
    )
  )

pbp <- pbp %>%
  group_by(Match_id) %>%
  tidyr::fill(pos, .direction = "down") %>%
  ungroup()

#loc
pbp <- pbp %>%
  mutate(
    loc = case_when(
      #Quarter start
      new_period ~ "M",
      
      # After score
      Stat == "Goal"   ~ "G",
      Stat == "Behind" ~ "B" , 
      
      #Trail after Goal
      lag(Stat) == "Goal" ~ "M",
      
      # Kick-ins
      Stat == "Kick In" & Team_status == "home" ~ "HK",
      Stat == "Kick In" & Team_status == "away" ~ "AK",
      
      # Inside 50 entries
      Stat == "Shot At Goal" & Team_status == "home" ~ "H50", 
      #Somehow in50s arent tracked????
      Stat == "Shot At Goal" & Team_status == "away" ~ "A50",
      # Inside 50 entries (marks)
      Stat == "F 50 Mark" & Team_status == "home" ~ "H50",
      Stat == "F 50 Mark" & Team_status == "away" ~ "A50",
      #F 50 Tackles
      Stat == "F 50 Tackle" & Team_status == "home" ~ "H50",
      Stat == "F 50 Tackle" & Team_status == "away" ~ "A50",
      
      #Inside 50 exits
      Stat == "Rebound 50" ~ "M",
      
      # Clearances → midfield proxy
      Stat %in% c("Centre Clearance") ~ "M",
      
      TRUE ~ NA_character_
    )
  )

pbp <- pbp %>%
  group_by(Match_id) %>%
  tidyr::fill(loc, .direction = "down") %>%
  ungroup()

# pos_id
pbp <- pbp %>%
  group_by(Match_id) %>%
  mutate(
    pos_start = row_number() == 1 |
                  pos != lag(pos),
    pos_id = cumsum(pos_start)
  ) %>%
  ungroup()

#infer from "rebound 50"

# That the current chain is relocated
pbp <- pbp %>%
  group_by(Match_id, pos_id) %>%
  mutate(row_in_chain = row_number()) %>%
  ungroup()

# Build rebound_info
rebound_info <- pbp %>%
  filter(Stat == "Rebound 50") %>%
  group_by(Match_id, pos_id) %>%
  summarise(
    rebound_row = min(row_in_chain),
    rebound_team = first(Team_status),
    .groups = "drop"
  )

# Join ONCE
pbp <- pbp %>%
  select(-any_of(c("rebound_row", "rebound_team"))) %>%
  left_join(rebound_info, by = c("Match_id","pos_id"))

# Now mutate safely
pbp <- pbp %>%
  mutate(
    loc = case_when(
      !is.na(rebound_row) &
      row_in_chain <= rebound_row &
      rebound_team == "home" &
      loc == "M" ~ "A50",

      !is.na(rebound_row) &
      row_in_chain <= rebound_row &
      rebound_team == "away" &
      loc == "M" ~ "H50",

      TRUE ~ loc
    )
  )

chain_rebound <- pbp %>%
  group_by(Match_id, pos_id) %>%
  summarise(
    has_rebound = any(Stat == "Rebound 50"),
    rebound_team = ifelse(
      has_rebound,
      first(Team_status[Stat == "Rebound 50"]),
      NA_character_
    ),
    .groups = "drop"
  )
chain_rebound <- chain_rebound %>%
  group_by(Match_id) %>%
  mutate(
    next_has_rebound = lead(has_rebound, default = FALSE),
    next_rebound_team = lead(rebound_team)
  ) %>%
  ungroup()
pbp <- pbp %>%
  left_join(
    chain_rebound %>%
      select(Match_id, pos_id, next_rebound_team),
    by = c("Match_id", "pos_id")
  )
pbp <- pbp %>%
  mutate(
    loc = case_when(
      !is.na(next_rebound_team) &
      pos == "Neutral" &
      next_rebound_team == "home" &
      loc == "M" ~ "A50",

      !is.na(next_rebound_team) &
      pos == "Neutral" &
      next_rebound_team == "away" &
      loc == "M" ~ "H50",

      TRUE ~ loc
    )
  )

#pos_and_loc
pbp <- pbp %>%
  mutate(
    pos_and_loc = case_when(
      
      #Quarter start
      new_period ~ "N_M",
      lag(Stat) == "Goal" ~ "N_M",
      
                  # After score
      Stat == "Goal"  & Team_status == "home" ~ "HG",
      Stat == "Goal"  & Team_status == "away" ~ "AG",
      Stat == "Behind"  & Team_status == "home" ~ "HB",
      Stat == "Behind"  & Team_status == "away" ~ "AB",
      
      # Kick-ins
      Stat == "Kick In" & Team_status == "home" ~ "HK",
      Stat == "Kick In" & Team_status == "away" ~ "AK",
      
      #H50 
      loc == "H50" & pos == "home" ~ "H_H50",
      loc == "H50" & (pos == "Neutral") ~ "N_H50",
      loc == "H50" & pos == "away" ~ "A_H50",
      
      #A50 
      loc == "A50" & pos == "home" ~ "H_A50",
      loc == "A50" & pos == "Neutral" ~ "N_A50",
      loc == "A50" & pos == "away" ~ "A_A50",
      
      #M
      loc == "M" & pos == "home" ~ "H_M",
      (loc == "M") & pos == "Neutral" ~ "N_M",
      (loc == "G") & pos == "Neutral" ~ "N_M", #Addressing G case
      loc == "M" & pos == "away" ~ "A_M",
      
      TRUE ~ NA_character_
    )
  )

pbp <- pbp %>%
  group_by(Match_id) %>%
  tidyr::fill(pos_and_loc, .direction = "down") %>%
  ungroup()

#state_change
pbp <- pbp %>%
  arrange(Match_id, Quarter, Time) %>%
  group_by(Match_id) %>%
  mutate(
    state_change = pos_and_loc != lag(pos_and_loc, default = first(pos_and_loc))
  ) %>%
  ungroup()
```

## Build Chain Starts2
```{r chain starts}
# Create chain_starts matrix
chain_starts <- pbp %>%
  filter(state_change)

chain_starts2 <- chain_starts %>%
  group_by(Match_id, Quarter) %>%
  mutate(
    next_pos_and_loc = lead(pos_and_loc)
  ) %>%
  ungroup() #%>%
  #filter(!is.na(next_pos_and_loc))

# for discrete soujorn times

chain_starts2<- chain_starts2%>%
  group_by(Match_id, Quarter) %>%
  mutate(
    pos_and_loc_time = lead(TimeSec) - TimeSec
  ) %>%
  ungroup()

# for individual game bayesian prediction model

chain_starts2 <- chain_starts2 %>%
  mutate(
    pos_and_loc_time_adj = if_else(
      ((startsWith(pos_and_loc, "N")) | (loc == "G") | (loc == "B") | is.na(pos_and_loc_time)),
      0,
      pos_and_loc_time
    )
  )

chain_starts2 <- chain_starts2 %>%
  group_by(Match_id) %>%
  mutate(
    match_time = cumsum(pos_and_loc_time_adj)
  ) %>%
  ungroup()

```

# Section 4.4

#Discrete soujourn times
```{r Discrete model soujourn times}
#We need to also have the distributions of time spent in each state for all of those that actually cause game time to elapse

discrete_chain_lengths <- chain_starts2%>%
  select(Match_id, Home, pos_id, pos_and_loc, next_pos_and_loc, pos_and_loc_time) %>%
  filter(pos_and_loc_time != 0) %>%
  filter(pos_and_loc_time < 150)

# Fit some gammas and exp to each of the ones that arent neutral
df_times <- discrete_chain_lengths %>%
  filter(pos_and_loc %in% c(
    "A_A50","A_M","A_H50",
    "H_A50","H_M","H_H50"
  ))

discrete_model_time_fits <- df_times %>%
  group_by(pos_and_loc) %>%
  summarise(
    exp_rate = fitdistr(pos_and_loc_time, "exponential")$estimate["rate"],
    gamma_shape = fitdistr(pos_and_loc_time, "gamma")$estimate["shape"],
    gamma_rate  = fitdistr(pos_and_loc_time, "gamma")$estimate["rate"]
  )

# Lets plot these fits to see how good they are:
densities <- discrete_model_time_fits %>%
  rowwise() %>%
  do({
    state <- .$pos_and_loc
    times <- seq(0, 150, length.out = 200)
    
    data.frame(
      pos_and_loc = state,
      time = times,
      exp_density = dexp(times, rate = .$exp_rate),
      gamma_density = dgamma(times, shape = .$gamma_shape, rate = .$gamma_rate)
    )
  })

p <- ggplot(df_times, aes(x = pos_and_loc_time)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "grey85",
    color = "grey30",
    linewidth = 0.2
  ) +
  geom_line(
    data = densities,
    aes(x = time, y = gamma_density, color = "Gamma Fit"),
    linewidth = 1
  ) +
  facet_wrap(
    ~ pos_and_loc,
    scales = "free_y"
  ) +
  coord_cartesian(xlim = c(0, 75)) +
  labs(
    title = "Holding Time Distributions by State",
    x = "Holding Time (seconds)",
    y = "Density",
    color = "Distribution"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14,
      hjust = 0.5
    ),
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 9),
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p

#______________________________________________________________________________
# Make a plot of with the next state included

p <- ggplot(
  df_times,
  aes(
    x = pos_and_loc_time,
    fill = next_pos_and_loc
  )
) +
  geom_bar() +
  facet_wrap(
    ~ pos_and_loc,
    scales = "free_y",
    ncol = 2
  ) +
  coord_cartesian(xlim = c(0, 75)) +
  labs(
    title = "Holding Times by Current and Next State",
    x = "Holding Time (seconds)",
    y = "Count",
    fill = "Next State"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14,
      hjust = 0.5
    ),
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 9),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p

```

# Mixture fits for A_A50 and H_H50
```{r gamma mixtures}

# Extract times for the attacking states
H_H50_times <- chain_starts2 %>%
  filter(
    pos_and_loc == "H_H50",
    pos_and_loc_time > 0,
    pos_and_loc_time < 70
  ) %>%
  pull(pos_and_loc_time)

A_A50_times <- chain_starts2 %>%
  filter(
    pos_and_loc == "A_A50",
    pos_and_loc_time > 0,
    pos_and_loc_time < 70
  ) %>%
  pull(pos_and_loc_time)

# Fit via EM Algorithm
H_H50_fit_mix <- gammamixEM(
  H_H50_times,
  k = 2,
  maxit = 5000,
  epsilon = 1e-8
)

cat("Gamma Params H_H50:")
H_H50_fit_mix$gamma.pars
cat("Weights H_H50:")
H_H50_fit_mix$lambda

A_A50_fit_mix <- gammamixEM(
  A_A50_times,
  k = 2,
  maxit = 5000,
  epsilon = 1e-8
)

cat("Gamma Params A_A50:")
A_A50_fit_mix$gamma.pars
cat("Weights A_A50:")
A_A50_fit_mix$lambda

# Function to generate mixture density data
get_gamma_mixture_density <- function(times, fit, state) {
  
  x <- seq(0, max(times), length.out = 1000)
  
  alpha1 <- fit$gamma.pars["alpha", "comp.1"]
  beta1  <- fit$gamma.pars["beta", "comp.1"]
  
  alpha2 <- fit$gamma.pars["alpha", "comp.2"]
  beta2  <- fit$gamma.pars["beta", "comp.2"]
  
  lambda1 <- fit$lambda[1]
  lambda2 <- fit$lambda[2]
  
  f1 <- dgamma(x, shape = alpha1, scale = beta1)
  f2 <- dgamma(x, shape = alpha2, scale = beta2)
  
  data.frame(
    time = x,
    component_1 = lambda1 * f1,
    component_2 = lambda2 * f2,
    mixture = lambda1 * f1 + lambda2 * f2,
    state = state
  )
}

H_H50_density <- get_gamma_mixture_density(
  H_H50_times,
  H_H50_fit_mix,
  "H_H50"
)

A_A50_density <- get_gamma_mixture_density(
  A_A50_times,
  A_A50_fit_mix,
  "A_A50"
)

mixture_densities <- rbind(
  H_H50_density,
  A_A50_density
)

mixture_times <- rbind(
  data.frame(
    time = H_H50_times,
    state = "H_H50"
  ),
  data.frame(
    time = A_A50_times,
    state = "A_A50"
  )
)

# Nice plot
p_mixture <- ggplot(mixture_times, aes(x = time)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 70,
    fill = "grey85",
    color = "grey30",
    linewidth = 0.2
  ) +
  geom_line(
    data = mixture_densities,
    aes(
      x = time,
      y = mixture,
      linetype = "Mixture"
    ),
    linewidth = 1,
    colour = "red"
  ) +
  facet_wrap(~ state, scales = "free_y") +
  coord_cartesian(xlim = c(0, 75)) +
  labs(
    title = "Gamma Mixture Fits for Selected Holding Time Distributions",
    x = "Holding Time (seconds)",
    y = "Density",
    linetype = "Distribution"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14,
      hjust = 0.5
    ),
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_mixture

```



# Section 5.1
## Build transition matrix

```{r transition matrix}

# Transition counts
transition_counts <- table(
  chain_starts2$pos_and_loc,
  chain_starts2$next_pos_and_loc
)

transition_probs3 <- prop.table(transition_counts, margin = 1)

# New order for the heatmap
new_order <- c(
  "HG", "HB", "AK",
  "H_H50", "N_H50", "A_H50",
  "H_M", "N_M", "A_M",
  "H_A50", "N_A50", "A_A50",
  "HK", "AB", "AG"
)

transition_probs3 <- transition_probs3[new_order, new_order]

# Heatmap
display_matrix <- matrix(
  sprintf("%.2f", transition_probs3),
  nrow = nrow(transition_probs3),
  dimnames = dimnames(transition_probs3)
)
display_matrix[transition_probs3 == 0] <- "0"

plot_data <- as.data.frame(as.table(transition_probs3))
colnames(plot_data) <- c("From", "To", "Probability")

# Force rows and columns to use the same ordering
state_order <- rownames(transition_probs3)

plot_data$From <- factor(plot_data$From, levels = rev(state_order))
plot_data$To <- factor(plot_data$To, levels = state_order)

# Labels for probabilities
plot_data$Label <- sprintf("%.2f", plot_data$Probability)
plot_data$Label[plot_data$Probability == 0] <- "0"

p <- ggplot(plot_data, aes(x = To, y = From, fill = Probability)) +
  geom_tile(colour = "grey80", linewidth = 0.3) +
  geom_text(aes(label = Label), size = 4) +
  scale_fill_gradientn(
    colours = c("white", "lightsteelblue1", "steelblue1")
  ) +
  scale_x_discrete(position = "bottom") +
  scale_y_discrete(position = "left") +
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
    panel.grid = element_blank()
  ) +
  labs(
    title = "Empirical AFL State Transition Matrix (Full 2021 Season Data)"
  )
p
```

## Sd of each transition game to game

We want to investigate how confident we are in this fit. How much it varies game to game.

```{r sd of transitions}

game_transitions <- chain_starts2 %>%
  filter(!is.na(next_pos_and_loc)) %>%
  count(Match_id, pos_and_loc, next_pos_and_loc) %>%
  group_by(Match_id, pos_and_loc) %>%
  mutate(p = n / sum(n)) %>%
  ungroup()

games <- unique(game_transitions$Match_id)
states <- sort(unique(c(game_transitions$pos_and_loc,
                        game_transitions$next_pos_and_loc)))

P_list <- lapply(games, function(g) {

  df_g <- game_transitions %>% filter(Match_id == g)

  mat <- matrix(0,
                nrow = length(states),
                ncol = length(states),
                dimnames = list(states, states))

  for (i in seq_len(nrow(df_g))) {
    mat[df_g$pos_and_loc[i], df_g$next_pos_and_loc[i]] <- df_g$p[i]
  }

  mat
})

names(P_list) <- games

P_array <- array(0,
                 dim = c(length(states), length(states), length(games)),
                 dimnames = list(states, states, games))

for (k in seq_along(games)) {
  P_array[,,k] <- P_list[[k]]
}

P_mean <- apply(P_array, c(1,2), mean, na.rm = TRUE)
P_sd <- apply(P_array, c(1,2), sd, na.rm = TRUE)

P_mean <- P_mean[new_order, new_order]
P_sd <- P_sd[new_order, new_order]

#new version of the plot for sds
plot_data <- as.data.frame(as.table(P_sd))
colnames(plot_data) <- c("From", "To", "SD")

# Force rows and columns to use the same ordering
state_order <- rownames(P_sd)

plot_data$From <- factor(plot_data$From, levels = rev(state_order))
plot_data$To <- factor(plot_data$To, levels = state_order)

# Labels for standard deviations
plot_data$Label <- sprintf("%.2f", plot_data$SD)
plot_data$Label[plot_data$SD == 0] <- "0"

p <- ggplot(plot_data, aes(x = To, y = From, fill = SD)) +
  geom_tile(colour = "grey80", linewidth = 0.3) +
  geom_text(aes(label = Label), size = 4) +
  scale_fill_gradientn(
    colours = c("white", "lightsteelblue1", "steelblue1")
  ) +
  scale_x_discrete(position = "bottom") +
  scale_y_discrete(position = "left") +
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
    panel.grid = element_blank()
  ) +
  labs(
    title = "Standard Deviation of Transition Probabilities over Matches"
  )

p

```

# Section 5.2
## Test the Markov assumption
```{r test the markov assumption}
# Prepare the data
# Create triples (Previous, Current, Next) within each
# Match_id and Quarter so that no transitions cross quarters.

test_data <- chain_starts2 %>%
  group_by(Match_id, Quarter) %>%
  mutate(
    Previous = lag(pos_and_loc, 1),
    Current = pos_and_loc,
    Next = lead(pos_and_loc, 1)
  ) %>%
  ungroup() %>%
  filter(
    !is.na(Previous),
    !is.na(Next)
  ) %>%
  select(Match_id, Quarter, Previous, Current, Next)


# Make sure the variables are factors
test_data <- test_data %>%
  mutate(
    Previous = factor(Previous),
    Current = factor(Current),
    Next = factor(Next)
  )

# ============================================================
# First-order multinomial model
model_1 <- multinom(
  Next ~ Current,
  data = test_data,
  trace = FALSE
)

# ============================================================
# Second-order multinomial model
model_2 <- multinom(
  Next ~ Previous + Current,
  data = test_data,
  trace = FALSE
)

# ============================================================
# Likelihood-ratio test
# Compare the log-likelihoods of the two models.

logLik(model_1)
logLik(model_2)

G2_reg <- 2 * (
  as.numeric(logLik(model_2)) -
  as.numeric(logLik(model_1))
)

G2_reg


# Degrees of freedom = difference in number of free parameters

df_reg <- attr(logLik(model_2), "df") -
          attr(logLik(model_1), "df")

df_reg


# Chi-squared p-value

p_value_reg <- pchisq(
  G2_reg,
  df = df_reg,
  lower.tail = FALSE
)

p_value_reg

# ============================================================
# Display the main test results

cat("First-order model parameters:", attr(logLik(model_1), "df"), "\n")
cat("Second-order model parameters:", attr(logLik(model_2), "df"), "\n")
cat("Degrees of freedom:", df_reg, "\n")
cat("Likelihood-ratio statistic:", G2_reg, "\n")
cat("p-value:", p_value_reg, "\n")

```

## Version with fixed zeroes 
```{r version with fixed zeroes}
#=============================================================
# First-order empirical probabilities

first_order <- test_data %>%
  count(Current, Next, name = "N_ij") %>%
  group_by(Current) %>%
  mutate(
    N_i = sum(N_ij),
    p_ij = N_ij / N_i
  ) %>%
  ungroup()


# Second-order empirical probabilities

second_order <- test_data %>%
  count(Previous, Current, Next, name = "N_kij") %>%
  group_by(Previous, Current) %>%
  mutate(
    N_ki = sum(N_kij),
    p_kij = N_kij / N_ki
  ) %>%
  ungroup()


df_first <- first_order %>%
  group_by(Current) %>%
  summarise(
    n_next = n(),
    .groups = "drop"
  ) %>%
  summarise(
    df = sum(n_next - 1)
  ) %>%
  pull(df)


df_second <- second_order %>%
  group_by(Previous, Current) %>%
  summarise(
    n_next = n(),
    .groups = "drop"
  ) %>%
  summarise(
    df = sum(n_next - 1)
  ) %>%
  pull(df)


df_direct <- df_second - df_first

# Match the first-order probabilities to the
# corresponding second-order transitions.

lr_data <- second_order %>%
  left_join(
    first_order %>%
      select(Current, Next, p_ij),
    by = c("Current", "Next")
  )


# Direct likelihood-ratio statistic

G2_direct <- 2 * sum(
  lr_data$N_kij *
    log(lr_data$p_kij / lr_data$p_ij)
)

G2_direct


p_value_direct <- pchisq(
  G2_direct,
  df = df_direct,
  lower.tail = FALSE
)
p_value_direct

cat("Direct first-order parameters:", df_first, "\n")
cat("Direct second-order parameters:", df_second, "\n")

```


## Find relative improvement
```{r relative improvement}
set.seed(123)

matches <- unique(test_data$Match_id)

train_matches <- sample(
  matches,
  size = 0.8 * length(matches)
)

train_data <- test_data %>%
  filter(Match_id %in% train_matches)

test_data_cv <- test_data %>%
  filter(!Match_id %in% train_matches)

model_1_cv <- multinom(
  Next ~ Current,
  data = train_data,
  trace = FALSE
)

model_2_cv <- multinom(
  Next ~ Previous + Current,
  data = train_data,
  trace = FALSE
)

pred_1 <- predict(
  model_1_cv,
  test_data_cv,
  type = "probs"
)

pred_2 <- predict(
  model_2_cv,
  test_data_cv,
  type = "probs"
)

p1 <- pred_1[
  cbind(
    seq_len(nrow(test_data_cv)),
    match(test_data_cv$Next, colnames(pred_1))
  )
]

p2 <- pred_2[
  cbind(
    seq_len(nrow(test_data_cv)),
    match(test_data_cv$Next, colnames(pred_2))
  )
]

log_loss_1 <- -mean(log(p1))
log_loss_2 <- -mean(log(p2))

log_loss_improvement <- log_loss_1 - log_loss_2

relative_improvement <- (
  log_loss_improvement / log_loss_1
) * 100

log_loss_1
log_loss_2
log_loss_improvement
relative_improvement

exp(log_loss_improvement)*100

```

# Section 5.3/ 5.4

## Long run behaviour of transition matrix
```{r limiting behaviour of discrete model}

#Split P into fast and slow states
P <- transition_probs3

# identify fast and slow states
fast_states <- c("AB", "AG", "HB", "HG", "N_M", "N_H50", "N_A50", "AK", "HK")
slow_states <- setdiff(rownames(P), fast_states)

# partition P
P_ss <- P[slow_states, slow_states]
P_sf <- P[slow_states, fast_states]
P_fs <- P[fast_states, slow_states]
P_ff <- P[fast_states, fast_states]

# reduced transition matrix
P_reduced <- P_ss + P_sf %*% solve(diag(length(fast_states)) - P_ff) %*% P_fs

#round(P_reduced, 4)

#Long run probabilites
A <- t(P) - diag(nrow(P))
A[nrow(A), ] <- 1   # replace last equation with sum constraint

b <- rep(0, nrow(P))
b[length(b)] <- 1

pi <- solve(A, b)

print('Here is the steady state:')
round(pi,4)

pi_full <- pi

# New plot:
steady_state_df <- tibble(
  state = rownames(P),
  pi = as.numeric(pi)
) %>%
  mutate(
    possession = case_when(
    str_detect(state, "^N") ~ "Neutral",
    str_detect(state, "^H") ~ "Home",
    str_detect(state, "^A") ~ "Away",
    TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(pi)) %>%
  mutate(state = factor(state, levels = state))

steady_state_plot <- ggplot(steady_state_df, aes(x = pi, y = state, fill = possession)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(pi, accuracy = 0.1)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c(
      "Neutral" = "forestgreen",
      "Home" = "firebrick",
      "Away" = "steelblue",
      "Other" = "grey70"
    )
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Steady-state distribution of possession-location states",
    subtitle = "Long-run proportion of transitions expected in each state",
    x = "Steady-state probability",
    y = NULL,
    fill = "Possession"
  ) +
  theme_minimal()

steady_state_plot

P <- P_reduced

A <- t(P) - diag(nrow(P))
A[nrow(A), ] <- 1   # replace last equation with sum constraint

b <- rep(0, nrow(P))
b[length(b)] <- 1

pi <- solve(A, b)

gamma_means <- with(discrete_model_time_fits, gamma_shape / gamma_rate)
names(gamma_means) <- discrete_model_time_fits$pos_and_loc

# update the means for the mixture states
A_A50_mean <- sum(
  A_A50_fit_mix$lambda *
  A_A50_fit_mix$gamma.pars["alpha", ] *
  A_A50_fit_mix$gamma.pars["beta", ]
)

H_H50_mean <- sum(
  H_H50_fit_mix$lambda *
  H_H50_fit_mix$gamma.pars["alpha", ] *
  H_H50_fit_mix$gamma.pars["beta", ]
)

gamma_means["A_A50"] <- A_A50_mean
gamma_means["H_H50"] <- H_H50_mean


gamma_means_full <- gamma_means[names(pi)]
gamma_means_full[is.na(gamma_means_full)] <- 0

weighted_times <- gamma_means_full * pi

prop_times <- weighted_times / sum(weighted_times)
print("Proportion of time in each state:")
round(prop_times,3)

prop_times_df <- tibble(
  state = names(prop_times),
  prop_time = as.numeric(prop_times)
) %>%
  mutate(
    possession = case_when(
      str_detect(state, "^N") ~ "Neutral",
      str_detect(state, "^H") ~ "Home",
      str_detect(state, "^A") ~ "Away",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(prop_time)) %>%
  mutate(state = factor(state, levels = state))

prop_time_plot <- ggplot(
  prop_times_df,
  aes(x = prop_time, y = state, fill = possession)
) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(prop_time, accuracy = 0.1)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c(
      "Neutral" = "forestgreen",
      "Home" = "firebrick",
      "Away" = "steelblue",
      "Other" = "grey70"
    )
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Proportion of time spent in each state",
    x = "Proportion of total time",
    y = NULL,
    fill = "Possession"
  ) +
  theme_minimal()

prop_time_plot

# make the ordering match lol
gamma_means_ordered <- gamma_means[names(prop_times)]

game_time <- 4 * 20 * 60

# Visit rates for slow states
visit_rate_slow <- prop_times / gamma_means_ordered

# Compute first-passage visits matrix F
F <- solve(diag(length(fast_states)) - P_ff) %*% P_fs

# Expected visits to fast states
expected_visits_slow <- visit_rate_slow * 4* 20 * 60 # your slow states
expected_visits_fast <- as.vector(expected_visits_slow %*% P_sf %*% solve(diag(length(fast_states)) - P_ff))
names(expected_visits_fast) <- fast_states

# Visit rates for fast states
visit_rate_fast <- expected_visits_fast / game_time

# Combine
visit_rate <- c(visit_rate_slow, visit_rate_fast)

# Put states in the same order as P
visit_rate <- visit_rate[rownames(transition_probs3)]

print("Visit rate:")
round(visit_rate, 5)

visit_rate_df <- tibble(
  state = names(visit_rate),
  visit_rate = as.numeric(visit_rate)
) %>%
  mutate(
    possession = case_when(
      str_detect(state, "^N") ~ "Neutral",
      str_detect(state, "^H") ~ "Home",
      str_detect(state, "^A") ~ "Away",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(visit_rate)) %>%
  mutate(state = factor(state, levels = state))

visit_rate_plot <- ggplot(
  visit_rate_df,
  aes(x = visit_rate, y = state, fill = possession)
) +
  geom_col() +
  geom_text(
    aes(label = round(visit_rate, 4)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c(
      "Neutral" = "forestgreen",
      "Home" = "firebrick",
      "Away" = "steelblue",
      "Other" = "grey70"
    )
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Visit rate by possession-location state",
    x = "Visit rate (visits per second)",
    y = NULL,
    fill = "Possession"
  ) +
  theme_minimal()

visit_rate_plot


names(expected_visits_fast) <- fast_states
print("Expected visits:")
expected_visits <- c(
  expected_visits_slow,
  expected_visits_fast
)

expected_visits <- expected_visits[rownames(transition_probs3)]

round(expected_visits, 2)

expected_visits_df <- tibble(
  state = names(expected_visits),
  expected_visits = as.numeric(expected_visits)
) %>%
  mutate(
    possession = case_when(
      str_detect(state, "^N") ~ "Neutral",
      str_detect(state, "^H") ~ "Home",
      str_detect(state, "^A") ~ "Away",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(expected_visits)) %>%
  mutate(state = factor(state, levels = state))

expected_visits_plot <- ggplot(
  expected_visits_df,
  aes(x = expected_visits, y = state, fill = possession)
) +
  geom_col() +
  geom_text(
    aes(label = round(expected_visits, 1)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_fill_manual(
    values = c(
      "Neutral" = "forestgreen",
      "Home" = "firebrick",
      "Away" = "steelblue",
      "Other" = "grey70"
    )
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Expected number of visits by possession-location state",
    x = "Expected number of visits per game",
    y = NULL,
    fill = "Possession"
  ) +
  theme_minimal()

expected_visits_plot

(expected_score <- c(expected_visits_fast["HG"]*6+ expected_visits_fast["HB"],
                     expected_visits_fast["AG"]*6+expected_visits_fast["AB"]))

```

##Gamma_params for pooled data simulation
```{r gamma params update}

gamma_params_wmix <- list()

gamma_params_wmix$A_A50 <- list(
  type = "gamma_mixture",
  alpha = A_A50_fit_mix$gamma.pars["alpha", ],
  beta = A_A50_fit_mix$gamma.pars["beta", ],
  lambda = A_A50_fit_mix$lambda
)

gamma_params_wmix$H_H50 <- list(
  type = "gamma_mixture",
  alpha = H_H50_fit_mix$gamma.pars["alpha", ],
  beta = H_H50_fit_mix$gamma.pars["beta", ],
  lambda = H_H50_fit_mix$lambda
)

gamma_params_wmix$A_H50 <- list(
  type = "gamma",
  shape = discrete_model_time_fits$gamma_shape[
    discrete_model_time_fits$pos_and_loc == "A_H50"
  ],
  rate = discrete_model_time_fits$gamma_rate[
    discrete_model_time_fits$pos_and_loc == "A_H50"
  ]
)

gamma_params_wmix$A_M <- list(
  type = "gamma",
  shape = discrete_model_time_fits$gamma_shape[
    discrete_model_time_fits$pos_and_loc == "A_M"
  ],
  rate = discrete_model_time_fits$gamma_rate[
    discrete_model_time_fits$pos_and_loc == "A_M"
  ]
)

gamma_params_wmix$H_A50 <- list(
  type = "gamma",
  shape = discrete_model_time_fits$gamma_shape[
    discrete_model_time_fits$pos_and_loc == "H_A50"
  ],
  rate = discrete_model_time_fits$gamma_rate[
    discrete_model_time_fits$pos_and_loc == "H_A50"
  ]
)

gamma_params_wmix$H_M <- list(
  type = "gamma",
  shape = discrete_model_time_fits$gamma_shape[
    discrete_model_time_fits$pos_and_loc == "H_M"
  ],
  rate = discrete_model_time_fits$gamma_rate[
    discrete_model_time_fits$pos_and_loc == "H_M"
  ]
)

gamma_params_wmix



```

## Pooled data simulation function
```{r simulate_discrete_gamma}
simulate_discrete_gamma <- function(P_list, gamma_params) {
  
  current_state <- "N_M"
  home <- 0
  away <- 0
  t <- 0
  quarter_time <- 20*60
  
  # Initialize counters
  total_times <- setNames(rep(0, length(names(P_list))), names(P_list))
  visit_counts <- setNames(rep(0, length(names(P_list))), names(P_list))
  
  states <- c(current_state)
  times <- c(0)
  home_score <- c(0)
  away_score <- c(0)
  
  for (q in 1:4){
    
    current_state <- "N_M"
    t <- 0
    
    while (t < quarter_time) {
      
      params <- gamma_params[[current_state]]

      # if rate is NA or 0, force dt = 0
      if(is.null(params)){
        dt <- 0
      } else if (params$type == "gamma") {
        dt <- rgamma(1,shape = params$shape, rate = params$rate)
      } else if (params$type == "gamma_mixture") {
        component <- sample(
          1:length(params$lambda),
          size = 1,
          prob = params$lambda
        )
        dt <- rgamma(
          1,
          shape = params$alpha[component],
          scale = params$beta[component]
        )
      } else {
        # BIG ERROR
        stop("Unknown distribution type: ", params$type)
      }
      
             # Track time and visits 
      if(t + dt < quarter_time){
        total_times[current_state] <- total_times[current_state] + dt
      }else {
        total_times[current_state] <- total_times[current_state] + (quarter_time - t)
      }
      
      visit_counts[current_state] <- visit_counts[current_state] + 1
      
      #Update score
      if (current_state == "HG"){
        home <- home + 6
      }
      if (current_state == "HB"){
        home <- home + 1
      }
      if (current_state == "AG"){
        away <- away + 6
      }
      if (current_state == "AB"){
        away <- away + 1
      }
      
      
      t <- t + dt
      if (t > quarter_time) break
      
      # sample next state
      next_info <- P_list[[current_state]]
      next_state <- sample(next_info$next_states, size = 1, prob = next_info$probs)
      
      current_state <- next_state
    }
  
  }
  
  return(
  data.frame(
    pos_and_loc = names(P_list),
    total_time = total_times,
    visits = visit_counts
  
  )
  )
}


```

##Pooled data model simulations
```{r Discrete model simulation}

set.seed(1223)

# Make sure rownames are the current states
states <- rownames(transition_probs3)

# convert transition matrix into list
P_list <- lapply(states, function(st) {
  probs <- transition_probs3[st, ]
  next_states <- names(probs)
  prob_vals <- as.numeric(probs)
  
  # remove 0-probabilities
  next_states <- next_states[prob_vals > 0]
  prob_vals <- prob_vals[prob_vals > 0]
  
  list(next_states = next_states, probs = prob_vals)
})
names(P_list) <- states

#Make a gamma model

gamma_params <- discrete_model_time_fits %>%
  dplyr::select(pos_and_loc, gamma_shape, gamma_rate) %>%
  split(.$pos_and_loc) %>%
  lapply(function(row) {
    list(
      shape = row$gamma_shape,
      rate  = row$gamma_rate
    )
  })

# Run multiple simulations and average
n <- 50000
sim_results <- replicate(n, simulate_discrete_gamma(P_list, gamma_params_wmix), simplify=FALSE)
sim_results_df <- bind_rows(sim_results, .id = "sim_id")

# Average per state across simulations
sim_summary <- sim_results_df %>%
  group_by(pos_and_loc) %>%
  summarise(
    mean_total_time = mean(total_time),
    mean_visits = mean(visits),
    sd_of_mean_visits = sd(visits)/sqrt(n),
    .groups = "drop"
  )

sim_summary
score <- c(sim_summary$mean_visits[7]+sim_summary$mean_visits[8]*6,
           sim_summary$mean_visits[1]+sim_summary$mean_visits[2]*6)
score
 
# Compare to the actual values 79.8 and 75.4

```

# Section 5.5
## Investigate rate of convergence of transition matrix to steady state

```{r rate of convergence of transition matrix}

P <- transition_probs3
#Compute Eigenvalues
eigens <- eigen(transition_probs3)
abs(eigens$values)

# Second largest eigenvalue
(rho <- abs(eigens$values[2]))
#(rho <- sort(Mod(eigens$values), decreasing = TRUE)[2])

# Diagonalise P
V <- eigens$vectors
Lambda <- diag(eigens$values)
V_inv <- solve(V)


# Derivation of the proper feller bound
# Identify the eigenvalue indices corresponding to 1 and 0
one_idx <- which.min(abs(eigens$values - 1))
zero_idx <- which.min(abs(eigens$values))

# Eigenvalues contributing to the convergence terms
convergence_idx <- setdiff(
  seq_along(eigens$values),
  c(one_idx, zero_idx)
)

# Construct b^(v) matrices
B <- lapply(seq_along(eigens$values), function(v) {
  outer(V[, v], V_inv[v, ])
})

# C_jk = sum_v |b_jk^(v)|, excluding lambda = 1 and lambda = 0
C_jk <- Reduce(
  `+`,
  lapply(convergence_idx, function(v) abs(B[[v]]))
)

# Feller constant
C_Feller <- max(rowSums(C_jk))

B_one <- outer(V[, one_idx], V_inv[one_idx, ])

print("Constant C:")
C_Feller

initial_state <- rep(0, 15)
initial_state[8] <- 1

num_iterations <- 30

error <- rep(NA,num_iterations)

for(i in 1:num_iterations){
  Pn <- diag(15)
  for(j in 1:i){
    Pn <- Pn %*% P
  }

  error[i] <- norm(initial_state %*% Pn - pi_full, type = "I")
}

# ============================================================
# Compute values for Fill bounds

P <- transition_probs3
pi <- pi_full

# D_pi
D_pi <- diag(pi)

# Adjoint of P in L^2(pi)
P_star <- solve(D_pi) %*% t(P) %*% D_pi

# Multiplicative reversibilisation
M <- P %*% P_star

# Eigenvalues of P*P
eigens_M <- eigen(M)
lambda_M <- sort(Re(eigens_M$values), decreasing = TRUE)

# Second-largest eigenvalue of P*P
beta2 <- lambda_M[2]

print("Sqrt(Beta_M):")
sqrt(beta2)

# pi_min
pi_min <- min(pi)

print("sqrt(Pi_min)")
sqrt(pi_min)

# Fill's theoretical bound
fill_bound <- beta2^(1:num_iterations/2) / sqrt(pi_min)

#=============================================================================
# New Fill bound in terms of A(P)

A <- (P + P_star) / 2

eigen_A <- eigen(A)

beta_a <- sort(Re(eigen_A$values), decreasing = TRUE)[2]

print("Sqrt(Beta_A):")
sqrt(beta_a)

fill_bound_A <- beta_a^(1:num_iterations/2) / sqrt(pi_min)


#=====================================================================

# Create data frame for plotting
bound_data <- data.frame(
  n = 1:num_iterations,
  Empirical = error,
  Feller = C_Feller * rho^(1:num_iterations),
  Fill_M = fill_bound,
  Fill_A = fill_bound_A
)

# Theoretical bounds in long format
bound_data_long <- reshape(
  bound_data,
  varying = c("Feller", "Fill_M", "Fill_A"),
  v.names = "value",
  timevar = "bound",
  times = c("Feller", "Fill_M", "Fill_A"),
  direction = "long"
)

# Empirical data
empirical_data <- data.frame(
  n = 1:num_iterations,
  value = error,
  bound = "Empirical"
)

# Combine all data
plot_data <- rbind(
  bound_data_long[, c("n", "value", "bound")],
  empirical_data
)

# Plot
p_boundcomparison <- ggplot() +

  # Theoretical bounds
  geom_line(
    data = bound_data_long,
    aes(x = n, y = value, colour = bound),
    linewidth = 0.9
  ) +

  # Empirical values
  geom_point(
    data = empirical_data,
    aes(x = n, y = value, colour = bound),
    size = 2
  ) +

  scale_y_log10(
    limits = c(1e-10, 1e3)
  ) +

  scale_colour_manual(
    values = c(
      "Empirical" = "black",
      "Feller" = "red",
      "Fill_M" = "blue",
      "Fill_A" = "darkgreen"
    ),
    labels = c(
    "Empirical" = "Empirical",
    "Feller" = "Feller bound",
    "Fill_M" = "Fill M bound",
    "Fill_A" = "Fill A bound"
  )
  ) +

  labs(
    title = "Convergence to the Stationary Distribution",
    x = "Number of transitions",
    y = "Infinity-norm error",
    colour = NULL
  ) +

  theme_classic(base_size = 12) +

  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

p_boundcomparison

```

# Section 6.1

## Get pooled match prior helper function
```{r get pooled match prior}
get_pooled_match_prior <- function(chain_starts, home_team, away_team, exclude = NA) {
  
  pooled_data <- chain_starts %>%
    filter(
      (Home == home_team | Away == away_team)
    )
  
  if (!is.na(exclude)) {
  pooled_data <- pooled_data %>%
    filter(Match_id != exclude)
  }
  
  # Transition probabilities
  transition_counts <- table(
    pooled_data$pos_and_loc,
    pooled_data$next_pos_and_loc
  )
  
  P <- prop.table(transition_counts, margin = 1)
  
  # States with holding times
  valid_states <- c(
    "A_A50", "A_M", "A_H50",
    "H_A50", "H_M", "H_H50"
  )
  
  time_data <- pooled_data %>%
    filter(
      pos_and_loc %in% valid_states,
      pos_and_loc_time > 0
    )
  
  # Ordinary gamma states
  gamma_states <- c(
    "A_H50", "A_M",
    "H_A50", "H_M"
  )
  gamma_params <- list()
  
  for (state in gamma_states) {
    
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- fitdistr(x, "gamma")
    
    gamma_params[[state]] <- list(
      type = "gamma",
      shape = unname(fit$estimate["shape"]),
      rate = unname(fit$estimate["rate"])
    )
  }
  # Mixture gamma states
  mixture_states <- c("A_A50", "H_H50")
  
  for (state in mixture_states) {
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- mixtools::gammamixEM(x,k = 2, maxit= 5000, verb = FALSE)
    
    gamma_params[[state]] <- list(
      type = "gamma_mixture",
      alpha = fit$gamma.pars["alpha", ],
      beta = fit$gamma.pars["beta", ],
      lambda = fit$lambda
    )
  }
  
  return(list(
    P = P,
    gamma = gamma_params
  ))
}
```

## Truncated simulator helper function
```{r truncated gamma sim}
#Truncated version of simulator for forecast
truncated_gamma_discrete_sim <- function(P_list, gamma_params, init_state, time, quarter, home, away, scale = FALSE){
  
  #Initial values
  current_state <- init_state
  t <- time
  quarter_time <- 20*60 #constant
  home_start <- home
  away_start <- away
  first_state <- TRUE
  
  #faster gamma 
  states <- names(gamma_params)
  state_index <- setNames(seq_along(states), states)
  
  #Finish current quarter
  while (t < quarter_time) {
      params <- gamma_params[[current_state]]

      # if rate is NA or 0, force dt = 0
      if(is.null(params)){
        dt <- 0
      } else if (params$type == "gamma") {
        dt <- rgamma(1,shape = params$shape, rate = params$rate)
      } else if (params$type == "gamma_mixture") {
        component <- sample(
          1:length(params$lambda),
          size = 1,
          prob = params$lambda
        )
        dt <- rgamma(
          1,
          shape = params$alpha[component],
          scale = params$beta[component]
        )
      } else {
        # BIG ERROR
        stop("Unknown distribution type: ", params$type)
      }
  
      
      if (!first_state){
        #Update score
        if (current_state == "HG"){
          home <- home + 6
        }
        if (current_state == "HB"){
          home <- home + 1
        }
        if (current_state == "AG"){
          away <- away + 6
        }
        if (current_state == "AB"){
          away <- away + 1
        }
      }
        first_state <- FALSE
        
      t <- t + dt
      if (t > quarter_time) break
      
      # sample next state
      next_info <- P_list[[current_state]]
      next_state <- sample(next_info$next_states, size = 1, prob = next_info$probs)
      
      
      current_state <- next_state
    }
  
  
  #Finish remaining quarters
  if (quarter < 4){
    for (q in (quarter+1):4){
      
      current_state <- "N_M"
      t <- 0
      
      while (t < quarter_time) {
        params <- gamma_params[[current_state]]
  
      # if rate is NA, force dt = 0
      if(is.null(params)){
        dt <- 0
      } else if (params$type == "gamma") {
        dt <- rgamma(1,shape = params$shape, rate = params$rate)
      } else if (params$type == "gamma_mixture") {
        component <- sample(
          1:length(params$lambda),
          size = 1,
          prob = params$lambda
        )
        dt <- rgamma(
          1,
          shape = params$alpha[component],
          scale = params$beta[component]
        )
      } else {
        # BIG ERROR
        stop("Unknown distribution type: ", params$type)
      }
  
        #Update score
        if (current_state == "HG"){
          home <- home + 6
        }
        if (current_state == "HB"){
          home <- home + 1
        }
        if (current_state == "AG"){
          away <- away + 6
        }
        if (current_state == "AB"){
          away <- away + 1
        }
          
        t <- t + dt
        if (t > quarter_time) break
        
        # sample next state
        next_info <- P_list[[current_state]]
        next_state <- sample(next_info$next_states, size = 1, prob = next_info$probs)
        
        current_state <- next_state
      }
    
    }
  }
  
  if (scale == TRUE){
    home <- home_start +
        scale_factor * (home - home_start)
    away <- away_start +
        scale_factor * (away - away_start)
  }
  
  forecast <- list(
  home_forecast = home,
  away_forecast = away
)
  return(forecast)
}

```


## P conv helper function
```{r Pconv}

P_conv <- function(transition_probs) {
  # Make sure rownames are the current states
  states <- rownames(transition_probs)
  
  P_list <- lapply(states, function(st) {
  probs <- transition_probs[st, ]
  next_states <- names(probs)
  prob_vals <- as.numeric(probs)
  
  # remove 0-probabilities
  next_states <- next_states[prob_vals > 0]
  prob_vals <- prob_vals[prob_vals > 0]
  
  list(next_states = next_states, probs = prob_vals)
  })
  names(P_list) <- states
  return(P_list)
}

```

## Individual game Bayesian Prediciton model
```{r build bayesian prediciton model}
#Forecaster

bayesian_score_forecast <- function(pbp,prior,nreps = 10,incr = 1, alpha = 0){
  
  # Priors
  prior_P <- prior$P
  P <- prior_P
  
  prior_gamma <- prior$gamma
  
  #Lets assume that we have already filtered the pbp to be the chain starts
  
  #Create a vector of match times
  cs_quarter_lengths <- tapply(pbp$match_time, pbp$Quarter, function(x) max(x, na.rm = TRUE))
  time <- pbp$match_time
  match_length <-  max(time, na.rm = TRUE)
  
  #Empty vectors for scores and predicted scores
  n <- length(time)
  home_scores <- rep(0,n)
  away_scores <- rep(0,n)
  home_forecast_scores <- rep(0,n)
  away_forecast_scores <- rep(0,n)
  home_win_rate <- rep(0,n)
  
  #Score counters
  home <- 0
  away <- 0
  
  #Loop through the entire pbp
  for (i in 1:n){
    if (i %% 50 == 0){cat("i =", i, "/", n, "\n")}
    pbprow <- pbp[i,]
    state <- pbprow$pos_and_loc
    next_state <- pbprow$next_pos_and_loc
    
    #Update real score if necessary
    if (state == "HG"){
      home <- home + 6
    }
    if (state == "HB"){
      home <- home + 1
    }
    if (state == "AG"){
      away <- away + 6
    }
    if (state == "AB"){
      away <- away + 1
    }
    
    #record actual scores
    home_scores[i] <- home
    away_scores[i] <- away
    
    #Update gamma params
    
    #Update P matrix
    P[state, ] <- (1 - alpha) * P[state, ]
    P[state, next_state] <- P[state, next_state] + alpha
    

    #Find "adjusted time"
    quarter <- pbp$Quarter[i]
    time_adjusted <- ifelse(quarter>1, (time[i] - cs_quarter_lengths[quarter-1]) / (cs_quarter_lengths[quarter] - cs_quarter_lengths[quarter-1]) * 1200, time[i] / cs_quarter_lengths[1] * 1200) 
    #Run simulations to forecast score
    forecast_sum <- c(0,0)
    home_wins <- 0
    
    if ((i %% incr == 0) | (i == 1) | (state %in% c("HG", "HB", "AG", "AB"))){
    for (k in 1:nreps){
        forecast <- truncated_gamma_discrete_sim(P_conv(P), prior_gamma, state, time_adjusted, quarter, home, away)
        forecast_sum <- forecast_sum + c(forecast$home_forecast, forecast$away_forecast)
        if (forecast$home_forecast > forecast$away_forecast){
          home_wins <- home_wins + 1
        }
      }
  
      home_forecast_scores[i] <- forecast_sum[1]/nreps
      away_forecast_scores[i] <- forecast_sum[2]/nreps
      home_win_rate[i] <- home_wins / nreps
    } else {
      home_forecast_scores[i] <- home_forecast_scores[i-1]
      away_forecast_scores[i] <- away_forecast_scores[i-1]
      home_win_rate[i] <- home_win_rate[i-1]
    }
    
    
  }
  
  # Create df for return
  score_forecast <- data.frame(
  Time = time,
  Home = home_scores,  
  Away = away_scores,
  Home_Forecast = home_forecast_scores,
  Away_Forecast = away_forecast_scores,
  Home_Win_Rate = home_win_rate
  ) 
  
  return(score_forecast)
}

bayesian_test_r1richvscarl <- chain_starts2[chain_starts2$Match_id == "CD_M20210140101",]
bayesian_test_r20gcvmelb <- chain_starts2[chain_starts2$Match_id == "CD_M20210142002",]

#Remove last row as it is problematic
bayesian_test_r20gcvmelb <- bayesian_test_r20gcvmelb %>%
  filter(!is.na(next_pos_and_loc))
#Remove last row as it is problematic
bayesian_test_r1richvscarl <- bayesian_test_r1richvscarl %>%
  filter(!is.na(next_pos_and_loc))

priors <- get_pooled_match_prior(chain_starts2, "Richmond", "Carlton", NA)
#priors <- get_pooled_match_prior(chain_start2, "Gold Coast Suns", "Melbourne")

increment <- 5

score_forecast <- bayesian_score_forecast(bayesian_test_r1richvscarl,priors,1000,increment,0)
#score_forecast <- bayesian_score_forecast(bayesian_test_r20gcvmelb,priors,250,5,0.01)

score_forecast <- score_forecast %>%
  mutate(
    Score_Diff = Home - Away
  )

score_forecast_long <- score_forecast %>%
  pivot_longer(
    cols = -Time,
    names_to = "Series",
    values_to = "Score"
  )

# Richmond vs carlton forecast plot
p <- ggplot(
  score_forecast_long %>%
    filter(Series %in% c(
      "Home", "Away",
      "Home_Forecast", "Away_Forecast"
    )),
  aes(x = Time, y = Score, colour = Series, linetype = Series)
) +
  geom_step(
    data = ~ subset(.x, Series %in% c("Home", "Away")),
    linewidth = 0.9
  ) +
  geom_step(
    data = ~ subset(.x, Series %in% c("Home_Forecast", "Away_Forecast")),
    linewidth = 0.9,
    alpha = 0.45
  ) +
  scale_colour_manual(
    values = c(
      "Home" = "#C44E52",
      "Away" = "#4C72B0",
      "Home_Forecast" = "#C44E52",
      "Away_Forecast" = "#4C72B0"
    ),
    labels = c(
      "Home" = "Home",
      "Away" = "Away",
      "Home_Forecast" = "Home forecast",
      "Away_Forecast" = "Away forecast"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "Home" = "solid",
      "Away" = "solid",
      "Home_Forecast" = "solid",
      "Away_Forecast" = "solid"
    ),
    labels = c(
      "Home" = "Home",
      "Away" = "Away",
      "Home_Forecast" = "Home forecast",
      "Away_Forecast" = "Away forecast"
    )
  ) +
  labs(
    title = "Cumulative Scores and Forecasted Final Scores",
    x = "Match time (s)",
    y = "Score",
    colour = NULL,
    linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p

# Win % plot
p <- ggplot(
  score_forecast_long %>%
    filter(Series %in% c("Score_Diff", "Home_Win_Rate")),
  aes(x = Time)
) +
  geom_hline(
  yintercept = 0,
  linetype = "dashed",
  linewidth = 0.5,
  colour = "grey50"
) +
  # Score difference
  geom_line(
    data = ~ subset(.x, Series == "Score_Diff"),
    aes(y = Score, colour = "Score difference"),
    linewidth = 1
  ) +
  
  # Home win probability
  geom_line(
    data = ~ subset(.x, Series == "Home_Win_Rate"),
    aes(y = (Score - 0.5) * 100, colour = "Home win probability"),
    linewidth = 1,
    alpha = 0.6
  ) +
  
  # Primary axis: score difference
  scale_y_continuous(
    name = "Score difference",
    sec.axis = sec_axis(
      ~ . / 100 + 0.5,
      name = "Home win probability",
      labels = scales::percent
    )
  ) +
  
  scale_colour_manual(
    values = c(
      "Score difference" = "black",
      "Home win probability" = "#4C72B0"
    )
  ) +
  
  labs(  title = "Score Difference and Home Win Probability",
    x = "Match time (s)",
    colour = NULL
  ) + theme(
  plot.title = element_text(
    hjust = 0.5,
    face = "bold"
  ),
  legend.position = "bottom",
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank()
) +
  
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.y.left = element_text(),
    axis.title.y.right = element_text(),
    legend.margin = margin(t = -5)
  ) +
  scale_y_continuous(
  name = "Score difference",
  limits = c(-50, 50),
  sec.axis = sec_axis(
    ~ . / 100 + 0.5,
    name = "Home win probability",
    labels = scales::percent
  )
)

p

```

## Selective version of bayesian prediction model
If we are going to simulate all games, need a version that only simulates every now and then
```{r selective version of bayesian prediction model}
bayesian_score_forecast_selective <- function(pbp,prior,nreps = 10, alpha = 0.01, scale = FALSE){
  
  # Priors
  prior_P <- prior$P
  P <- prior_P
  
  prior_gamma <- prior$gamma

  #Lets assume that we have already filtered the pbp to be the chain starts
  
  #Create a vector of match times
  cs_quarter_lengths <- tapply(pbp$match_time, pbp$Quarter, function(x) max(x, na.rm = TRUE))
  time <- pbp$match_time
  match_length <-  max(time, na.rm = TRUE)
  
  #Empty vectors for scores and predicted scores
  n <- length(time)
quarter_forecasts <- data.frame(
  Quarter = integer(),
  Home_Score = integer(),
  Away_Score = integer(),
  Home_Forecast = numeric(),
  Away_Forecast = numeric(),
  Home_Win_Rate = numeric()
)
  
  #Score counters
  home <- 0
  away <- 0
  
  #Quater ends
  quarter_end <- c(
  which(pbp$Quarter[-nrow(pbp)] != pbp$Quarter[-1]),
  nrow(pbp)
)
  
  #Loop through the entire pbp
  for (i in 1:n){
    pbprow <- pbp[i,]
    state <- pbprow$pos_and_loc
    next_state <- pbprow$next_pos_and_loc
    
    #Update real score if necessary
    if (state == "HG"){
      home <- home + 6
    }
    if (state == "HB"){
      home <- home + 1
    }
    if (state == "AG"){
      away <- away + 6
    }
    if (state == "AB"){
      away <- away + 1
    }
    
    #Update P matrix
    P[state, ] <- (1 - alpha) * P[state, ]
    P[state, next_state] <- P[state, next_state] + alpha
    

    #Find "adjusted time"
    quarter <- pbp$Quarter[i]
    time_adjusted <- ifelse(quarter>1, (time[i] - cs_quarter_lengths[quarter-1]) / (cs_quarter_lengths[quarter] - cs_quarter_lengths[quarter-1]) * 1200, time[i] / cs_quarter_lengths[1] * 1200) 
    #Run simulations to forecast score
    forecast_sum <- c(0,0)
    home_wins <- 0
    
if (i %in% quarter_end){

    forecast_sum <- c(0,0)
    home_wins <- 0

    for (k in 1:nreps){

        forecast <- truncated_gamma_discrete_sim(
          P_conv(P),
          prior_gamma,
          state,
          time_adjusted,
          quarter,
          home,
          away,
          scale
        )

        forecast_sum <- forecast_sum +
          c(forecast$home_forecast,
            forecast$away_forecast)

        if (forecast$home_forecast > forecast$away_forecast){
          home_wins <- home_wins + 1
        }
    }

    quarter_forecasts <- rbind(
      quarter_forecasts,
      data.frame(
        Quarter = quarter,
        Home_Score = home,
        Away_Score = away,
        Home_Forecast = forecast_sum[1]/nreps,
        Away_Forecast = forecast_sum[2]/nreps,
        Home_Win_Rate = home_wins/nreps
      )
    )
}
    
  }
  
  return(quarter_forecasts)
}

```


## Simulate error for the null and EWMA models
We want to compare the predicted final score at QT,HT,3QT with the actual score and find the error. This can be done for the whole season.
```{r}

# Firstly lets see 
season_simulator <- function(chain_starts,nreps,alpha,scale=FALSE){
  all_results <- list()         
  
  counter <- 0
  for (match in unique(chain_starts$Match_id)) {
    if(counter %% 10 == 1){cat("i =", counter, "/", 22*9, "\n")}
    # Split target match
    test_match <- chain_starts %>%
      filter(Match_id == match)
    
    test_match <- test_match %>%
  filter(!is.na(next_pos_and_loc))
    
    # Extract teams and score
    home_team <- unique(test_match$Home)
    away_team <- unique(test_match$Away)
    
    # Build prior excluding this match
    priors <- get_pooled_match_prior(chain_starts, home_team, 
                                     away_team, exclude = match)
    
    # Run Bayesian forecast
    forecast <- bayesian_score_forecast_selective(
      test_match,
      priors,
      nreps,
      alpha,
      scale
    )
    
    final_home <- forecast$Home_Score[forecast$Quarter == 4]
    final_away <- forecast$Away_Score[forecast$Quarter == 4]
    
    # Compute error
    forecast <- forecast %>%
  mutate(
    Match_id = match,
    Home_Team = unique(test_match$Home),
    Away_Team = unique(test_match$Away),
    Final_Home = final_home,
    Final_Away = final_away,
    Home_Error = abs(Home_Forecast - final_home),
    Away_Error = abs(Away_Forecast - final_away),
    Score_MAE = (Home_Error + Away_Error)/2
  )
    
    # Store
    all_results[[match]] <- forecast
    
    
    counter <- counter + 1;
  }
  
  results <- bind_rows(all_results)
  return(results)
}

# Each of these takes about half an hour
season_forecasts2_null <- suppressWarnings(season_simulator(chain_starts2,1000,0,FALSE))
season_forecasts2_ewma_alpha0001 <- suppressWarnings(season_simulator(chain_starts2,1000,0.001,FALSE))
season_forecasts2_ewma_alpha0002 <- suppressWarnings(season_simulator(chain_starts2,1000,0.002,FALSE))
season_forecasts2_ewma_alpha0005 <- suppressWarnings(season_simulator(chain_starts2,1000,0.005,FALSE))
season_forecasts2_ewma_alpha001 <- suppressWarnings(season_simulator(chain_starts2,1000,0.01,FALSE))
season_forecasts2_ewma_alpha002 <- suppressWarnings(season_simulator(chain_starts2,1000,0.02,FALSE))
season_forecasts2_ewma_alpha005 <- suppressWarnings(season_simulator(chain_starts2,1000,0.05,FALSE))
season_forecasts2_ewma_alpha01 <- suppressWarnings(season_simulator(chain_starts2,1000,0.1,FALSE))

 season_forecasts2_ewma_alpha0003 <- suppressWarnings(season_simulator(chain_starts2,1000,0.003,FALSE))
 season_forecasts2_ewma_alpha0007 <- suppressWarnings(season_simulator(chain_starts2,1000,0.007,FALSE))
 season_forecasts2_ewma_alpha0004 <- suppressWarnings(season_simulator(chain_starts2,1000,0.004,FALSE))
 season_forecasts2_ewma_alpha0006 <- suppressWarnings(season_simulator(chain_starts2,1000,0.006,FALSE))
 season_forecasts2_ewma_alpha0008 <- suppressWarnings(season_simulator(chain_starts2,1000,0.008,FALSE))


```



## Create a extrapolation model for comparison
Take the score at QT*2, HT*2 and 3QT* 4/3
```{r extrapolation model}

forecast_extrapolate <- function(pbp){
  
  #Create a vector of match times
  cs_quarter_lengths <- tapply(pbp$match_time, pbp$Quarter, function(x) max(x, na.rm = TRUE))
  time <- pbp$match_time
  match_length <-  max(time, na.rm = TRUE)
  
  #Empty vectors for scores and predicted scores
  n <- length(time)
  quarter_forecasts <- data.frame(
    Quarter = integer(),
    Home_Score = integer(),
    Away_Score = integer(),
    Home_Forecast = numeric(),
    Away_Forecast = numeric(),
    Home_Win_Rate = numeric()
  )
  
  #Score counters
  home <- 0
  away <- 0
  
  #Quater ends
  quarter_end <- c(which(pbp$Quarter[-nrow(pbp)] != pbp$Quarter[-1]), nrow(pbp))
  
  #Loop through the entire pbp
  for (i in 1:n){
    pbprow <- pbp[i,]
    state <- pbprow$pos_and_loc
    next_state <- pbprow$next_pos_and_loc
    
    #Update real score if necessary
    if (state == "HG"){
      home <- home + 6
    }
    if (state == "HB"){
      home <- home + 1
    }
    if (state == "AG"){
      away <- away + 6
    }
    if (state == "AB"){
      away <- away + 1
    }

    if (i %in% quarter_end){
    
    quarter <- pbp$Quarter[i]  
      
    if (quarter == 1){
      home_forecast <- home * 4
      away_forecast <- away * 4
    } else if(quarter == 2){
      home_forecast <- home * 2
      away_forecast <- away * 2 
    } else if(quarter == 3){
      home_forecast <- home * 4/3
      away_forecast <- away * 4/3
    } else{
      home_forecast <- home
      away_forecast <- away
    }
      
    #Update gamma params
    quarter_forecasts <- rbind(
      quarter_forecasts,
      data.frame(
        Quarter = quarter,
        Home_Score = home,
        Away_Score = away,
        Home_Forecast = home_forecast,
        Away_Forecast = away_forecast
      )
    )
  
    }
  }
  return(quarter_forecasts)
}

#----------------------------------------------------------------------------

season_simulator_extrapolate <- function(chain_starts){
  all_results <- list() 
  counter <- 1
  for (match in unique(chain_starts$Match_id)) {
      if(counter %% 10 == 1){cat("i =", counter, "/", 22*9, "\n")}
      # Split target match
      test_match <- chain_starts %>%
        filter(Match_id == match)
      
    #   test_match <- test_match %>%
    # filter(!is.na(next_pos_and_loc))
      
      # Extract teams and score
      home_team <- unique(test_match$Home)
      away_team <- unique(test_match$Away)
      
      # Extrapolate
      forecast <- forecast_extrapolate(test_match)
      
      final_home <- forecast$Home_Score[forecast$Quarter == 4]
      final_away <- forecast$Away_Score[forecast$Quarter == 4]
      
      # Compute error
      forecast <- forecast %>%
    mutate(
      Match_id = match,
      Home_Team = unique(test_match$Home),
      Away_Team = unique(test_match$Away),
      Final_Home = final_home,
      Final_Away = final_away,
      Home_Error = (Home_Forecast - final_home),
      Away_Error = (Away_Forecast - final_away)
    )
      
      # Store
      all_results[[match]] <- forecast
      
      
      counter <- counter + 1;
    }
    
    results <- bind_rows(all_results)
    return(results)
  }

#_____________________________________________________________________________
season_forecasts_extrapolations <- season_simulator_extrapolate(chain_starts2)

forecast_extrapolations_errors <- season_forecasts_extrapolations %>%
  filter(Quarter < 4)%>%
  mutate(
    Home_Abs_Error = abs(Home_Forecast - Final_Home),
    Away_Abs_Error = abs(Away_Forecast - Final_Away),
    Total_Abs_Error = abs(Home_Forecast - Final_Home) + abs(Away_Forecast - Final_Away)
  )

```


# Section 6.2

## Get match prior counts helper function
```{r Get pooled match prior counts}
get_pooled_match_prior_counts <- function(chain_starts, home_team, away_team, exclude = NA) {
  
  pooled_data <- chain_starts %>%
    filter(
      (Home == home_team | Away == away_team)
    )
  
  if (!is.na(exclude)) {
  pooled_data <- pooled_data %>%
    filter(Match_id != exclude)
  }
  
  # Transition probabilities
  transition_counts <- table(
    pooled_data$pos_and_loc,
    pooled_data$next_pos_and_loc
  )
  
  P_counts <- transition_counts
  
  # States with holding times
  valid_states <- c(
    "A_A50", "A_M", "A_H50",
    "H_A50", "H_M", "H_H50"
  )
  
  time_data <- pooled_data %>%
    filter(
      pos_and_loc %in% valid_states,
      pos_and_loc_time > 0
    )
  
  # ---------------------------------------------------
  # Ordinary gamma states
  
  gamma_states <- c(
    "A_H50", "A_M",
    "H_A50", "H_M"
  )
  
  gamma_params <- list()
  
  for (state in gamma_states) {
    
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- fitdistr(x, "gamma")
    
    gamma_params[[state]] <- list(
      type = "gamma",
      shape = unname(fit$estimate["shape"]),
      rate = unname(fit$estimate["rate"])
    )
  }
  
  
  # ----------------------------------------------------------------
  # Mixture gamma states
  
  mixture_states <- c("A_A50", "H_H50")
  
  for (state in mixture_states) {
    
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- mixtools::gammamixEM(x,k = 2, maxit= 5000, verb = FALSE)
    
    gamma_params[[state]] <- list(
      type = "gamma_mixture",
      alpha = fit$gamma.pars["alpha", ],
      beta = fit$gamma.pars["beta", ],
      lambda = fit$lambda
    )
  }
  
  return(list(
    P_counts = P_counts,
    gamma = gamma_params
  ))
}

```


## Selective Dirichlet model (bayesian_score_forecast_selective_dirchlet)
```{r Dirichlet prediction model}
# So the first thing that we have to do is now store the P_counts as inputs 
# rather than the P_probs so that we can update them
bayesian_score_forecast_selective_dirchlet <- function(pbp,prior_counts,
                                                       nreps = 10, factor = 1, 
                                                       scale = FALSE){
  
  # Priors
  P_counts <- prior_counts$P_counts
  
  prior_gamma <- prior_counts$gamma
  
  #Lets assume that we have already filtered the pbp to be the chain starts
  
  #Create a vector of match times
  cs_quarter_lengths <- tapply(pbp$match_time, pbp$Quarter, function(x) max(x, na.rm = TRUE))
  time <- pbp$match_time
  match_length <-  max(time, na.rm = TRUE)
  
  #Empty vectors for scores and predicted scores
  n <- length(time)
quarter_forecasts <- data.frame(
  Quarter = integer(),
  Home_Score = integer(),
  Away_Score = integer(),
  Home_Forecast = numeric(),
  Away_Forecast = numeric(),
  Home_Win_Rate = numeric()
)
  
  #Score counters
  home <- 0
  away <- 0
  
  #Quater ends
  quarter_end <- c(
  which(pbp$Quarter[-nrow(pbp)] != pbp$Quarter[-1]),
  nrow(pbp)
)
  
  #Loop through the entire pbp
  for (i in 1:n){
    pbprow <- pbp[i,]
    state <- pbprow$pos_and_loc
    next_state <- pbprow$next_pos_and_loc
    
    #Update real score if necessary
    if (state == "HG"){
      home <- home + 6
    }
    if (state == "HB"){
      home <- home + 1
    }
    if (state == "AG"){
      away <- away + 6
    }
    if (state == "AB"){
      away <- away + 1
    }

    #Update gamma params
    
    #Update P_counts matrix
    P_counts[state, next_state] <- P_counts[state, next_state] + 1*factor
    P <- prop.table(P_counts, margin = 1)

    #Find "adjusted time"
    quarter <- pbp$Quarter[i]
    time_adjusted <- ifelse(quarter>1, (time[i] - cs_quarter_lengths[quarter-1]) / (cs_quarter_lengths[quarter] - cs_quarter_lengths[quarter-1]) * 1200, time[i] / cs_quarter_lengths[1] * 1200) 
    #Run simulations to forecast score
    forecast_sum <- c(0,0)
    home_wins <- 0
    
if (i %in% quarter_end){

    forecast_sum <- c(0,0)
    home_wins <- 0

    for (k in 1:nreps){
        forecast <- truncated_gamma_discrete_sim(
          P_conv(P),
          prior_gamma,
          state,
          time_adjusted,
          quarter,
          home,
          away,
          scale
        )
        forecast_sum <- forecast_sum +
          c(forecast$home_forecast,
            forecast$away_forecast)

        if (forecast$home_forecast > forecast$away_forecast){
          home_wins <- home_wins + 1
        }
    }
    quarter_forecasts <- rbind(
      quarter_forecasts,
      data.frame(
        Quarter = quarter,
        Home_Score = home,
        Away_Score = away,
        Home_Forecast = forecast_sum[1]/nreps,
        Away_Forecast = forecast_sum[2]/nreps,
        Home_Win_Rate = home_wins/nreps
      )
    )
}
    
  }
  
  return(quarter_forecasts)
}


```

##Test Dirichlet model
```{r test dirichlet model}
season_simulator_dirichlet <- function(chain_starts,nreps,factor = 1,scale=FALSE){
  all_results <- list()         
  
  counter <- 0
  for (match in unique(chain_starts$Match_id)) {
    if(counter %% 10 == 1){cat("i =", counter, "/", 22*9, "\n")}
    # Split target match
    test_match <- chain_starts %>%
      filter(Match_id == match)
    
    test_match <- test_match %>%
  filter(!is.na(next_pos_and_loc))
    
    # Extract teams and score
    home_team <- unique(test_match$Home)
    away_team <- unique(test_match$Away)
    
    # Build prior excluding this match
    priors <- get_pooled_match_prior_counts(chain_starts, home_team, 
                                     away_team, exclude = match)
    
    # Run Bayesian forecast
    forecast <- bayesian_score_forecast_selective_dirchlet(
      test_match,
      priors,
      nreps,
      factor,
      scale
    )
    
    final_home <- forecast$Home_Score[forecast$Quarter == 4]
    final_away <- forecast$Away_Score[forecast$Quarter == 4]
    
    # Compute error
    forecast <- forecast %>%
  mutate(
    Match_id = match,
    Home_Team = unique(test_match$Home),
    Away_Team = unique(test_match$Away),
    Final_Home = final_home,
    Final_Away = final_away,
    Home_Error = abs(Home_Forecast - final_home),
    Away_Error = abs(Away_Forecast - final_away),
    Score_MAE = (Home_Error + Away_Error)/2
  )
    
    # Store
    all_results[[match]] <- forecast
    
    
    counter <- counter + 1;
  }
  
  results <- bind_rows(all_results)
  return(results)
}

# This takes about half an hour
season_forecasts2_dirichlet <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 1, FALSE))
season_forecasts2_dirichlet2 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 2, FALSE))
season_forecasts2_dirichlet3 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 3, FALSE))
season_forecasts2_dirichlet4 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 4, FALSE))
season_forecasts2_dirichlet5 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 5, FALSE))
season_forecasts2_dirichlet10 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 10, FALSE))
season_forecasts2_dirichlet20 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 20, FALSE))
season_forecasts2_dirichlet50 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 50, FALSE))

 season_forecasts2_dirichlet7 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 7, FALSE))
 season_forecasts2_dirichlet6 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 6, FALSE))
 season_forecasts2_dirichlet8 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 8, FALSE))
 season_forecasts2_dirichlet9 <- suppressWarnings(season_simulator_dirichlet(chain_starts2,1000,factor = 9, FALSE))
```


# Section 6.3
## EMWA errors are in 6.1 section

# Section 6.4
## Average scores by quarter
```{r average score by quarter}
quarter_scores_long <- season_forecasts_extrapolations %>%
  select(Match_id, Quarter, Home_Score, Away_Score) %>%
  pivot_longer(
    cols = c(Home_Score, Away_Score),
    names_to = "Team",
    values_to = "Score"
  ) %>%
  mutate(
    Team = recode(
      Team,
      Home_Score = "Home",
      Away_Score = "Away"
    )
  )

quarter_scoring_summary <- quarter_scores_long %>%
  arrange(Match_id, Team, Quarter) %>%
  group_by(Match_id, Team) %>%
  mutate(
    Quarter_Score = Score - lag(Score, default = 0)
  ) %>%
  ungroup() %>%
  group_by(Quarter, Team) %>%
  summarise(
    Mean_Quarter_Score = mean(Quarter_Score, na.rm = TRUE),
    SD_Quarter_Score = sd(Quarter_Score, na.rm = TRUE),
    .groups = "drop"
  )

quarter_scoring_summary

# Plot average points scored during each quarter
p <- ggplot(
  quarter_scoring_summary,
  aes(
    x = Quarter,
    y = Mean_Quarter_Score,
    colour = Team,
    group = Team
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Average Points Scored During Each Quarter",
    x = "Quarter",
    y = "Average Points Scored",
    colour = NULL
  ) +
  scale_x_continuous(
    breaks = 1:4
  ) +
  scale_colour_manual(
    values = c(
      "Home" = "red",
      "Away" = "blue"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )
p

cumulative_scoring_summary <- quarter_scoring_summary %>%
  arrange(Team, Quarter) %>%
  group_by(Team) %>%
  mutate(
    Mean_Cumulative_Score = cumsum(Mean_Quarter_Score)
  ) %>%
  ungroup() %>%
  bind_rows(
    tibble(
      Team = c("Home", "Away"),
      Quarter = 0,
      Mean_Cumulative_Score = 0
    )
  ) %>%
  arrange(Team, Quarter)

p <- ggplot(
  cumulative_scoring_summary,
  aes(
    x = Quarter,
    y = Mean_Cumulative_Score,
    colour = Team,
    group = Team
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Average Cumulative Score After Each Quarter",
    x = "Quarter",
    y = "Average Cumulative Score",
    colour = NULL
  ) +
  scale_x_continuous(
    breaks = 0:4,
    labels = c("Start", "Q1", "Q2", "Q3", "Q4")
  ) +
  scale_colour_manual(
    values = c(
      "Home" = "red",
      "Away" = "blue"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )

p

```

## Variation between and within quarters
```{r varation within between quarters}
# Divide each quarter into 10 equal bins

quarter_lengths <- pbp %>%
  group_by(Match_id, Quarter) %>%
  summarise(
    quarter_length = max(TimeSec, na.rm = TRUE),
    .groups = "drop"
  )

bin_width <- 10

score_events <- pbp %>%
  filter(Stat %in% c("Goal", "Behind")) %>%
  left_join(
    quarter_lengths,
    by = c("Match_id", "Quarter")
  ) %>%
  mutate(
    quarter_pct = 100 * TimeSec / quarter_length,

    Team = case_when(
      Team_status == "home" ~ "Home",
      Team_status == "away" ~ "Away"
    ),

    Score = case_when(
      Stat == "Goal" ~ 6,
      Stat == "Behind" ~ 1
    ),

    bin = cut(
      quarter_pct,
      breaks = seq(0, 100, bin_width),
      include.lowest = TRUE,
      labels = FALSE
    )
  )

# ------------------------------------------------------------
# Score within each tenth of each quarter

bin_scores <- score_events %>%
  group_by(Match_id, Quarter, Team, bin) %>%
  summarise(
    score = sum(Score),
    .groups = "drop"
  ) %>%
  group_by(Match_id, Quarter, Team) %>%
  complete(
    bin = 1:10,
    fill = list(score = 0)
  ) %>%
  ungroup()

# ------------------------------------------------------------
# Mean score within each tenth of each quarter

mean_bin_scores <- bin_scores %>%
  group_by(Quarter, Team, bin) %>%
  summarise(
    mean_score = mean(score),
    .groups = "drop"
  ) %>%
  mutate(
    bin_midpoint = (bin - 0.5) * bin_width
  )

mean_bin_scores

quarter_scores <- score_events %>%
  group_by(Match_id, Quarter, Team) %>%
  summarise(
    score = sum(Score),
    .groups = "drop"
  ) %>%
  complete(
    Match_id,
    Quarter = 1:4,
    Team,
    fill = list(score = 0)
  ) %>%
  pivot_wider(
    names_from = Quarter,
    values_from = score,
    names_prefix = "Q"
  )

home_scores <- quarter_scores %>%
  filter(Team == "Home")

away_scores <- quarter_scores %>%
  filter(Team == "Away")

friedman.test(
  as.matrix(home_scores[, c("Q1", "Q2", "Q3", "Q4")])
)

friedman.test(
  as.matrix(away_scores[, c("Q1", "Q2", "Q3", "Q4")])
)


library(broom)

trend_tests <- mean_bin_scores %>%
  group_by(Team, Quarter) %>%
  do(
    tidy(lm(mean_score ~ bin, data = .))
  ) %>%
  ungroup() %>%
  filter(term == "bin") %>%
  select(Team, Quarter, estimate, std.error, statistic, p.value) %>%
  mutate(
    Quarter = paste0("Q", Quarter)
  )

trend_tests

p <- ggplot(
  mean_bin_scores,
  aes(
    x = bin,
    y = mean_score,
    colour = Team
  )
) +
  geom_point(size = 2) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  facet_grid(
    Team ~ Quarter,
    labeller = labeller(
      Quarter = function(x) paste0("Q", x)
    )
  ) +
  scale_x_continuous(
    breaks = 1:10
  ) +
  scale_colour_manual(
    values = c(
      "Home" = "red",
      "Away" = "blue"
    )
  ) +
  labs(
    title = "Within-quarter trends in mean scoring",
    x = "10% segment of quarter",
    y = "Mean score",
    colour = NULL
  ) +
  theme_minimal()

away_long <- away_scores %>%
  select(Match_id, Q1, Q2, Q3, Q4) %>%
  pivot_longer(
    cols = starts_with("Q"),
    names_to = "Quarter",
    values_to = "Score"
  )

wilcox.test(away_scores$Q1, away_scores$Q2, paired = TRUE)
wilcox.test(away_scores$Q2, away_scores$Q3, paired = TRUE)
wilcox.test(away_scores$Q3, away_scores$Q4, paired = TRUE)

wilcox.test(home_scores$Q1, home_scores$Q2, paired = TRUE)
wilcox.test(home_scores$Q2, home_scores$Q3, paired = TRUE)
wilcox.test(home_scores$Q3, home_scores$Q4, paired = TRUE)

away_quarter_long <- away_scores %>%
  pivot_longer(
    cols = Q1:Q4,
    names_to = "Quarter",
    values_to = "Score"
  )

ggplot(
  away_quarter_long,
  aes(x = Score)
) +
  geom_histogram(
    binwidth = 5,
    boundary = 0,
    colour = "black"
  ) +
  geom_vline(
    data = away_quarter_long %>%
      group_by(Quarter) %>%
      summarise(
        Mean = mean(Score),
        .groups = "drop"
      ),
    aes(xintercept = Mean),
    linetype = "dashed",
    linewidth = 1
  ) +
  facet_wrap(~ Quarter, ncol = 2) +
  labs(
    title = "Distribution of Away-Team Scores by Quarter",
    x = "Score",
    y = "Number of Matches"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    panel.grid.minor = element_blank()
  )

away_quarter_long %>%
  group_by(Quarter) %>%
  summarise(
    Mean = mean(Score),
    Median = median(Score),
    Proportion_30_plus = mean(Score >= 30),
    Count_30_plus = sum(Score >= 30),
    Proportion_40_plus = mean(Score >= 40),
    Count_40_plus = sum(Score >= 40),
    Proportion_50_plus = mean(Score >= 50),
    Count_50_plus = sum(Score >= 50),
    n = n()
  )

```


## Transition matrix by quarter
```{r transiton matrix by quarter}
# Transition matrix fitted by quarter
# Function to estimate transition matrix for a given quarter
get_transition_matrix <- function(data, quarter) {
  
  counts <- data %>%
    filter(
      Quarter == quarter,
      !is.na(next_pos_and_loc)
    ) %>%
    count(pos_and_loc, next_pos_and_loc) %>%
    tidyr::pivot_wider(
      names_from = next_pos_and_loc,
      values_from = n,
      values_fill = 0
    ) %>%
    as.data.frame()
  
  # Convert to matrix
  states <- new_order
  
  transition_counts <- matrix(
    0,
    nrow = length(states),
    ncol = length(states),
    dimnames = list(states, states)
  )
  
  for (i in seq_len(nrow(counts))) {
    from <- counts$pos_and_loc[i]
    
    for (j in seq_along(states)) {
      to <- states[j]
      
      if (to %in% names(counts)) {
        transition_counts[from, to] <- counts[i, to]
      }
    }
  }
  
  # Convert counts to probabilities
  transition_probs <- prop.table(
    transition_counts,
    margin = 1
  )
  
  return(list(
    counts = transition_counts,
    probabilities = transition_probs
  ))
}


# Estimate a matrix for each quarter
P_Q1 <- get_transition_matrix(chain_starts2, 1)
P_Q2 <- get_transition_matrix(chain_starts2, 2)
P_Q3 <- get_transition_matrix(chain_starts2, 3)
P_Q4 <- get_transition_matrix(chain_starts2, 4)


# Put the four quarter matrices into a 3D array
P_array <- array(
  c(P_Q1$probabilities, P_Q2$probabilities, P_Q3$probabilities, P_Q4$probabilities),
  dim = c(nrow(P_Q1$probabilities), ncol(P_Q1$probabilities), 4)
)

P_array

# Standard deviation across quarters
sd_matrix <- apply(P_array, c(1, 2), sd)

# Make sure row/column names are retained
dimnames(sd_matrix) <- list(new_order, new_order)

#new version of this plot
plot_data <- as.data.frame(as.table(sd_matrix))
colnames(plot_data) <- c("From", "To", "SD")

# Convert SDs to percentage points
plot_data$SD <- plot_data$SD * 100

# Force rows and columns to use the same ordering
state_order <- rownames(sd_matrix)

plot_data$From <- factor(plot_data$From, levels = rev(state_order))
plot_data$To <- factor(plot_data$To, levels = state_order)

# Labels for standard deviations
plot_data$Label <- sprintf("%.2f", plot_data$SD)
plot_data$Label[plot_data$SD == 0] <- "0"

p_qbyqtrans_sd <- ggplot(
  plot_data,
  aes(x = To, y = From, fill = SD)
) +
  geom_tile(colour = "grey80", linewidth = 0.3) +
  geom_text(aes(label = Label), size = 4) +
  scale_fill_gradientn(
    colours = c("white", "lightsteelblue1", "steelblue1")
  ) +
  scale_x_discrete(position = "bottom") +
  scale_y_discrete(position = "left") +
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
    panel.grid = element_blank()
  ) +
  labs(
    title = "SD of Transition Probabilities Across Quarters (x10^-2)"
  )

p_qbyqtrans_sd

# Which transitions change the most qbyq
# Quarter-specific matrices
P_list <- list(
  Q1 = P_Q1$probabilities,
  Q2 = P_Q2$probabilities,
  Q3 = P_Q3$probabilities,
  Q4 = P_Q4$probabilities
)

# Convert a transition matrix to a long data frame
matrix_to_df <- function(P) {
  as.data.frame(P) %>%
    rownames_to_column("from") %>%
    pivot_longer(
      -from,
      names_to = "to",
      values_to = "probability"
    )
}

# Combine into one data frame
transition_long <- bind_rows(
  lapply(names(P_list), function(q) {
    matrix_to_df(P_list[[q]]) %>%
      mutate(Quarter = q)
  })
)

transition_changes <- transition_long %>%
  select(from, to, Quarter, probability) %>%
  pivot_wider(
    names_from = Quarter,
    values_from = probability
  ) %>%
  mutate(
    Q2_Q1 = Q2 - Q1,
    Q3_Q2 = Q3 - Q2,
    Q4_Q3 = Q4 - Q3,
    Q4_Q1 = Q4 - Q1,
    max_change = pmax(
      abs(Q2_Q1),
      abs(Q3_Q2),
      abs(Q4_Q3)
    )
  ) %>%
  arrange(desc(max_change))

transition_changes

```

## QbyQ version of truncated simulator helper function
```{r truncated gamma sim qbyq}
truncated_gamma_discrete_sim_qbyq <- function(P_Q1, P_Q2, P_Q3, P_Q4, gamma_params, init_state, time, quarter, home, away, scale = FALSE){
  
  #Initial values
  current_state <- init_state
  t <- time
  quarter_time <- 20*60 #constant
  home_start <- home
  away_start <- away
  first_state <- TRUE
  
  # Store transition matrices in a list
  P_lists <- list(
    P_Q1 = P_Q1,
    P_Q2 = P_Q2,
    P_Q3 = P_Q3,
    P_Q4 = P_Q4
  )
  
  # Set P for current Q
  P_list <- P_lists[[paste0("P_Q", quarter)]]
  
  #faster gamma 
  states <- names(gamma_params)
  state_index <- setNames(seq_along(states), states)
  
  
  #Finish current quarter
  while (t < quarter_time) {
      params <- gamma_params[[current_state]]

      # if rate is NA or 0, force dt = 0
      if(is.null(params)){
        dt <- 0
      } else if (params$type == "gamma") {
        dt <- rgamma(1,shape = params$shape, rate = params$rate)
      } else if (params$type == "gamma_mixture") {
        component <- sample(
          1:length(params$lambda),
          size = 1,
          prob = params$lambda
        )
        dt <- rgamma(
          1,
          shape = params$alpha[component],
          scale = params$beta[component]
        )
      } else {
        # BIG ERROR
        stop("Unknown distribution type: ", params$type)
      }
      
      if (!first_state){
        #Update score
        if (current_state == "HG"){
          home <- home + 6
        }
        if (current_state == "HB"){
          home <- home + 1
        }
        if (current_state == "AG"){
          away <- away + 6
        }
        if (current_state == "AB"){
          away <- away + 1
        }
      }
        first_state <- FALSE
        
      t <- t + dt
      if (t > quarter_time) break
      
      # sample next state
      next_info <- P_list[[current_state]]
      if (is.null(next_info) || length(next_info$next_states) == 0) {
      stop(
        "No valid transitions from state: ", current_state,
        " in quarter: ", q)
      }
      next_state <- sample(next_info$next_states, size = 1, 
                           prob = next_info$probs)
      
      
      current_state <- next_state
    }
  
  
  #Finish remaining quarters
  if (quarter < 4){
    for (q in (quarter+1):4){
      
      current_state <- "N_M"
      t <- 0
      
      P_list <- P_lists[[paste0("P_Q", q)]]
      
      while (t < quarter_time) {
        params <- gamma_params[[current_state]]
  
      # if rate is NA, force dt = 0
      if(is.null(params)){
        dt <- 0
      } else if (params$type == "gamma") {
        dt <- rgamma(1,shape = params$shape, rate = params$rate)
      } else if (params$type == "gamma_mixture") {
        component <- sample(
          1:length(params$lambda),
          size = 1,
          prob = params$lambda
        )
        dt <- rgamma(
          1,
          shape = params$alpha[component],
          scale = params$beta[component]
        )
      } else {
        # BIG ERROR
        stop("Unknown distribution type: ", params$type)
      }
  
        #Update score
        if (current_state == "HG"){
          home <- home + 6
        }
        if (current_state == "HB"){
          home <- home + 1
        }
        if (current_state == "AG"){
          away <- away + 6
        }
        if (current_state == "AB"){
          away <- away + 1
        }
          
        t <- t + dt
        if (t > quarter_time) break
        
        # sample next state
        next_info <- P_list[[current_state]]
        if (is.null(next_info) || length(next_info$next_states) == 0) {
  stop(
    "No valid transitions: state = ", current_state,
    ", quarter = ", q
  )
}
        next_state <- sample(next_info$next_states, size = 1, prob = next_info$probs)
        
        current_state <- next_state
      }
    
    }
  }
  
  if (scale == TRUE){
    home <- home_start +
        scale_factor * (home - home_start)
    away <- away_start +
        scale_factor * (away - away_start)
  }
  
  forecast <- list(
  home_forecast = home,
  away_forecast = away
)
  return(forecast)
}
```

## Qbyq version of selective simulator
```{r qbyq bayesian score forecast selective qbyq}

bayesian_score_forecast_selective_qbyq <- function(pbp,prior,nreps = 10, alpha = 0.01, scale = FALSE){
  
  # Priors
  P <- prior$P_Q1
  
    P_sim <- list(
      P_Q1 = prior$P_Q1,
      P_Q2 = prior$P_Q2,
      P_Q3 = prior$P_Q3,
      P_Q4 = prior$P_Q4
    )
  
  prior_gamma <- prior$gamma
  
  #Lets assume that we have already filtered the pbp to be the chain starts
  
  #Create a vector of match times
  cs_quarter_lengths <- tapply(pbp$match_time, pbp$Quarter, function(x) max(x, na.rm = TRUE))
  time <- pbp$match_time
  match_length <-  max(time, na.rm = TRUE)
  
  #Empty vectors for scores and predicted scores
  n <- length(time)
quarter_forecasts <- data.frame(
  Quarter = integer(),
  Home_Score = integer(),
  Away_Score = integer(),
  Home_Forecast = numeric(),
  Away_Forecast = numeric(),
  Home_Win_Rate = numeric()
)
  
  #Score counters
  home <- 0
  away <- 0
  
  #Quater ends
  quarter_end <- c(
  which(pbp$Quarter[-nrow(pbp)] != pbp$Quarter[-1]),
  nrow(pbp)
)
  
  #Loop through the entire pbp
  for (i in 1:n){
    pbprow <- pbp[i,]
    state <- pbprow$pos_and_loc
    next_state <- pbprow$next_pos_and_loc
    
    #Update real score if necessary
    if (state == "HG"){
      home <- home + 6
    }
    if (state == "HB"){
      home <- home + 1
    }
    if (state == "AG"){
      away <- away + 6
    }
    if (state == "AB"){
      away <- away + 1
    }

    #Update gamma params
    
    #Update P matrix
    P[state, ] <- (1 - alpha) * P[state, ]
    P[state, next_state] <- P[state, next_state] + alpha
    
    # ACTUALLY update all the priors based on what has occured
    P_sim$P_Q1[state, ] <- (1 - alpha) * P_sim$P_Q1[state, ]
    P_sim$P_Q1[state, next_state] <- P_sim$P_Q1[state, next_state] + alpha
    P_sim$P_Q2[state, ] <- (1 - alpha) * P_sim$P_Q2[state, ]
    P_sim$P_Q2[state, next_state] <- P_sim$P_Q2[state, next_state] + alpha
    P_sim$P_Q3[state, ] <- (1 - alpha) * P_sim$P_Q3[state, ]
    P_sim$P_Q3[state, next_state] <- P_sim$P_Q3[state, next_state] + alpha
    P_sim$P_Q4[state, ] <- (1 - alpha) * P_sim$P_Q4[state, ]
    P_sim$P_Q4[state, next_state] <- P_sim$P_Q4[state, next_state] + alpha
    
    
    #Find "adjusted time"
    quarter <- pbp$Quarter[i]
    time_adjusted <- ifelse(quarter>1, (time[i] - cs_quarter_lengths[quarter-1]) / (cs_quarter_lengths[quarter] - cs_quarter_lengths[quarter-1]) * 1200, time[i] / cs_quarter_lengths[1] * 1200) 
    #Run simulations to forecast score
    forecast_sum <- c(0,0)
    home_wins <- 0
    
if (i %in% quarter_end){

    forecast_sum <- c(0,0)
    home_wins <- 0
      
      # Replace the current quarter's prior with its
      # Bayesian-updated matrix
      P_sim[[paste0("P_Q", quarter)]] <- P
   
    for (k in 1:nreps){

        forecast <- truncated_gamma_discrete_sim_qbyq(
          P_conv(P_sim$P_Q1),P_conv(P_sim$P_Q2),P_conv(P_sim$P_Q3),
          P_conv(P_sim$P_Q4),
          prior_gamma,
          state,
          time_adjusted,
          quarter,
          home,
          away,
          scale
        )

        forecast_sum <- forecast_sum +
          c(forecast$home_forecast,
            forecast$away_forecast)

        if (forecast$home_forecast > forecast$away_forecast){
          home_wins <- home_wins + 1
        }
    }

    quarter_forecasts <- rbind(
      quarter_forecasts,
      data.frame(
        Quarter = quarter,
        Home_Score = home,
        Away_Score = away,
        Home_Forecast = forecast_sum[1]/nreps,
        Away_Forecast = forecast_sum[2]/nreps,
        Home_Win_Rate = home_wins/nreps
      )
    )
    
    # Change the p for new quarter
    if (quarter == 1){
      P <- prior$P_Q2
    } else if (quarter ==2 ){
      P <- prior$P_Q3
    } else if (quarter == 3){
      P <- prior$P_Q4
    }
}
    
  }
  
  return(quarter_forecasts)
}

```

## QbyQ match prior helper function
```{r pooled match prior for qbyq}
get_pooled_match_prior_qbyq <- function(chain_starts, home_team, 
                                     away_team, exclude = match) {
  
  pooled_data <- chain_starts %>%
    filter(
      (Home == home_team | Away == away_team)
    )
  
  if (!is.na(exclude)) {
  pooled_data <- pooled_data %>%
    filter(Match_id != exclude)
  }
  
    transition_states <- c(
    "HG", "HB", "AK",
    "H_H50", "N_H50", "A_H50",
    "H_M", "N_M", "A_M",
    "H_A50", "N_A50", "A_A50",
    "HK", "AB", "AG"
  )
  
  # Function to calculate P for one quarter
  get_P <- function(data, quarter) {
    
    quarter_data <- data %>%
      filter(
        Quarter == quarter,
        !is.na(next_pos_and_loc)
      )
    
    transition_counts <- table(
  factor(quarter_data$pos_and_loc, levels = transition_states),
  factor(quarter_data$next_pos_and_loc, levels = transition_states)
)
    
    P <- prop.table(transition_counts, margin = 1)
    
    # Rows with no observed transitions will contain NaN
    P[is.nan(P)] <- 0
    
    return(P)
  }
  
  # Get one transition matrix for each quarter
  P_Q1 <- get_P(pooled_data, 1)
  P_Q2 <- get_P(pooled_data, 2)
  P_Q3 <- get_P(pooled_data, 3)
  P_Q4 <- get_P(pooled_data, 4)
  
  # States with holding times
  valid_states <- c(
    "A_A50", "A_M", "A_H50",
    "H_A50", "H_M", "H_H50"
  )
  
  time_data <- pooled_data %>%
    filter(
      pos_and_loc %in% valid_states,
      pos_and_loc_time > 0
    )
  
  # -----------------------------
  # Ordinary gamma states
  # -----------------------------
  
  gamma_states <- c(
    "A_H50", "A_M",
    "H_A50", "H_M"
  )
  
  gamma_params <- list()
  
  for (state in gamma_states) {
    
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- fitdistr(x, "gamma")
    
    gamma_params[[state]] <- list(
      type = "gamma",
      shape = unname(fit$estimate["shape"]),
      rate = unname(fit$estimate["rate"])
    )
  }
  
  # -----------------------------
  # Mixture gamma states
  # -----------------------------
  
  mixture_states <- c("A_A50", "H_H50")
  
  for (state in mixture_states) {
    
    x <- time_data %>%
      filter(pos_and_loc == state) %>%
      pull(pos_and_loc_time)
    
    fit <- mixtools::gammamixEM(x,k = 2, maxit= 5000, verb = FALSE)
    
    gamma_params[[state]] <- list(
      type = "gamma_mixture",
      alpha = fit$gamma.pars["alpha", ],
      beta = fit$gamma.pars["beta", ],
      lambda = fit$lambda
    )
  }
  
  return(list(
    P_Q1 = P_Q1,
    P_Q2 = P_Q2,
    P_Q3 = P_Q3,
    P_Q4 = P_Q4,
    gamma = gamma_params
  ))
}
```


##Simulated qbyq errors
```{r qbyq model errors}
season_simulator_qbyq <- function(chain_starts,nreps,alpha,scale=FALSE){
  all_results <- list()         
  
  counter <- 0
  for (match in unique(chain_starts$Match_id)) {
    if(counter %% 10 == 1){cat("i =", counter, "/", 22*9, "\n")}
    # Split target match
    test_match <- chain_starts %>%
      filter(Match_id == match)
    
    test_match <- test_match %>%
  filter(!is.na(next_pos_and_loc))
    
    # Extract teams and score
    home_team <- unique(test_match$Home)
    away_team <- unique(test_match$Away)
    
    # Build prior excluding this match
    priors <- get_pooled_match_prior_qbyq(chain_starts, home_team, 
                                     away_team, exclude = match)
    
    # Run Bayesian forecast
    forecast <- bayesian_score_forecast_selective_qbyq(
      test_match,
      priors,
      nreps,
      alpha,
      scale
    )
    
    final_home <- forecast$Home_Score[forecast$Quarter == 4]
    final_away <- forecast$Away_Score[forecast$Quarter == 4]
    
    # Compute error
    forecast <- forecast %>%
  mutate(
    Match_id = match,
    Home_Team = unique(test_match$Home),
    Away_Team = unique(test_match$Away),
    Final_Home = final_home,
    Final_Away = final_away,
    Home_Error = abs(Home_Forecast - final_home),
    Away_Error = abs(Away_Forecast - final_away),
    Score_MAE = (Home_Error + Away_Error)/2
  )
    
    # Store
    all_results[[match]] <- forecast
    
    
    counter <- counter + 1;
  }
  
  results <- bind_rows(all_results)
  return(results)
}

# each takes a while like an hour or smth

season_forecasts2_qbyq_null <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0,FALSE))
season_forecasts2_qbyq_0001 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.001,FALSE))
season_forecasts2_qbyq_0002 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.002,FALSE))
season_forecasts2_qbyq_0005 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.005,FALSE))
season_forecasts2_qbyq_001 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.01,FALSE))
season_forecasts2_qbyq_002 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.02,FALSE))

 season_forecasts2_qbyq_0003 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.003,FALSE))
 season_forecasts2_qbyq_0007 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.007,FALSE))
 season_forecasts2_qbyq_0004 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.004,FALSE))
 season_forecasts2_qbyq_0006 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.006,FALSE))
 season_forecasts2_qbyq_0008 <- suppressWarnings(season_simulator_qbyq(chain_starts2,1000,0.008,FALSE))

```

# Section 7.4

## All models comparison table

```{r comparison errors}
# Calculate forecast errors

# Function to calculate forecast errors
calculate_forecast_errors <- function(x) {
  x %>%
    filter(Quarter < 4) %>%
    mutate(
      Home_Bias = Home_Forecast - Final_Home,
      Away_Bias = Away_Forecast - Final_Away,
      Home_Abs_Error = abs(Home_Forecast - Final_Home),
      Away_Abs_Error = abs(Away_Forecast - Final_Away),
      Total_Abs_Error = Home_Abs_Error + Away_Abs_Error
    )
}

# List of forecast objects
forecast_names <- c(
  "null",
  "dirichlet",
  "dirichlet10",
  "dirichlet5",
  "dirichlet50",
  "dirichlet20",
  "dirichlet2",
  "dirichlet3",
  "dirichlet4",
  "dirichlet6",
  "dirichlet7",
  "dirichlet8",
  "dirichlet9",
  "ewma_alpha001",
  "ewma_alpha002",
  "ewma_alpha005",
  "ewma_alpha01",
  "ewma_alpha0001",
  "ewma_alpha0002",
  "ewma_alpha0005",
  "ewma_alpha0003",
  "ewma_alpha0004",
  "ewma_alpha0006",
  "ewma_alpha0007",
  "ewma_alpha0008",
  "qbyq_null",
  "qbyq_0001",
  "qbyq_0002",
  "qbyq_0005",
  "qbyq_0003",
  "qbyq_0004",
  "qbyq_0006",
  "qbyq_0007",
  "qbyq_0008",
  "qbyq_001",
  "qbyq_002"
)


# Apply function to each forecast object
for (name in forecast_names) {
  forecast_object <- get(paste0("season_forecasts2_", name))
  assign(
    paste0("forecast_errors_", name),
    calculate_forecast_errors(forecast_object)
  )
}

# Map object suffixes to model labels
model_labels <- c(
  null = "Null model",
  extrapolations_errors = "Extrapolation model",
  
  ewma_alpha001 = "EWMA model 0.01",
  ewma_alpha002 = "EWMA model 0.02",
  ewma_alpha005 = "EWMA model 0.05",
  ewma_alpha01 = "EWMA model 0.1",
  ewma_alpha0001 = "EWMA model 0.001",
  ewma_alpha0002 = "EWMA model 0.002",
  ewma_alpha0005 = "EWMA model 0.005",
  ewma_alpha0003 = "EWMA model 0.003",
  ewma_alpha0004 = "EWMA model 0.004",
  ewma_alpha0006 = "EWMA model 0.006",
  ewma_alpha0007 = "EWMA model 0.007",
  ewma_alpha0008 = "EWMA model 0.008",
  
  dirichlet = "Dirichlet model",
  dirichlet2 = "Dirichlet model 2",
  dirichlet3 = "Dirichlet model 3",
  dirichlet4 = "Dirichlet model 4",
  dirichlet5 = "Dirichlet model 5",
  dirichlet6 = "Dirichlet model 6",
  dirichlet7 = "Dirichlet model 7",
  dirichlet8 = "Dirichlet model 8",
  dirichlet9 = "Dirichlet model 9",
  dirichlet10 = "Dirichlet model 10",
  dirichlet20 = "Dirichlet model 20",
  dirichlet50 = "Dirichlet model 50",
  
  qbyq_null = "QbyQ model 0",
  qbyq_001 = "QbyQ model 0.01",
  qbyq_002 = "QbyQ model 0.02",
  qbyq_0005 = "QbyQ model 0.005",
  qbyq_0006 = "QbyQ model 0.006",
  qbyq_0007 = "QbyQ model 0.007",
  qbyq_0008 = "QbyQ model 0.008",
  qbyq_0003 = "QbyQ model 0.003",
  qbyq_0004 = "QbyQ model 0.004",
  qbyq_0002 = "QbyQ model 0.002",
  qbyq_0001 = "QbyQ model 0.001"
)

# Combine all error data frames
comparison_errors <- bind_rows(
  lapply(names(model_labels), function(name) {
    
    # Extrapolation has a slightly different object name
        if (name == "extrapolations_errors") {
      data <- forecast_extrapolations_errors %>%
        mutate(
          Home_Bias = Home_Forecast - Final_Home,
          Away_Bias = Away_Forecast - Final_Away,
          Home_Abs_Error = abs(Home_Forecast - Final_Home),
          Away_Abs_Error = abs(Away_Forecast - Final_Away),
          Total_Abs_Error = Home_Abs_Error + Away_Abs_Error
        )
    } else {
      data <- get(paste0("forecast_errors_", name))
    }
    
    data %>%
      mutate(Model = model_labels[[name]])
  })
)

error_summary <- comparison_errors %>%
  select(Model, Quarter, Home_Bias, Away_Bias,
         Home_Abs_Error, Away_Abs_Error) %>%
  group_by(Model, Quarter) %>%
  summarise(
    Home_Bias = mean(Home_Bias),
    Away_Bias = mean(Away_Bias),
    Home_MAE = mean(Home_Abs_Error),
    Away_MAE = mean(Away_Abs_Error),
    .groups = "drop"
  )

error_summary

overall_summary <- error_summary %>%
  group_by(Model) %>%
  summarise(
    Home_Bias = mean(abs(Home_Bias)),
    Away_Bias = mean(abs(Away_Bias)),
    Home_MAE = mean(Home_MAE),
    Away_MAE = mean(Away_MAE),
    Overall_MAE = (mean(Home_MAE) + mean(Away_MAE))/2,
    .groups = "drop"
  ) %>% arrange(Overall_MAE)
overall_summary

```

## Compare each to null
```{r compare to null}
# Compare each model against the Null model

model_difference <- comparison_errors %>%
  select(Match_id, Quarter, Model, Total_Abs_Error) %>%
  pivot_wider(
    names_from = Model,
    values_from = Total_Abs_Error
  ) %>%
  pivot_longer(
    cols = -c(Match_id, Quarter, `Null model`),
    names_to = "Model",
    values_to = "Model_Error"
  ) %>%
  mutate(
    Difference = Model_Error - `Null model`
  )


# ---------------------------------------------------------
# Calculate mean difference and 95% CI

model_difference_summary <- model_difference %>%
  group_by(Model, Quarter) %>%
  summarise(
    Mean_Difference = mean(Difference, na.rm = TRUE),
    SD = sd(Difference, na.rm = TRUE),
    n = sum(!is.na(Difference)),
    SE = SD / sqrt(n),
    CI = qt(0.975, df = n - 1) * SE,
    CI_lower = Mean_Difference - CI,
    CI_upper = Mean_Difference + CI,
    .groups = "drop"
  )
model_difference_summary

```


## Null vs extrapolation plots ?
```{r null vs extrapolation}
p <- ggplot(
  comparison_errors %>%
    filter(Model %in% c("Null model", "Extrapolation model")),
  aes(x = Total_Abs_Error, fill = Model)
) +
  geom_histogram(
    bins = 30,
    alpha = 0.5,
    position = "identity"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(
  ~ factor(
    Quarter,
    levels = c(1, 2, 3),
    labels = c("QT", "HT", "3QT")
  )
) +
  labs(
    title = "Total Absolute Error Histograms by Quarter Break",
    x = "Total Absolute Error",
    y = "Count",
    fill = "Model"
  ) +
  theme_minimal()

p

```

## Null vs Dirichlet 1 plot
```{r null vs dirichlet 1}

p <- ggplot(
  comparison_errors %>%
    filter(Model %in% c("Null model", "Dirichlet model")),
  aes(x = Total_Abs_Error, fill = Model)
) +
  geom_histogram(
    bins = 30,
    alpha = 0.5,
    position = "identity"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(
  ~ factor(
    Quarter,
    levels = c(1, 2, 3),
    labels = c("QT", "HT", "3QT")
  )) +
  labs(
    title = "Total Absolute Error Histograms by Quarter Break",
    x = "Total Absolute Error",
    y = "Count",
    fill = "Model"
  ) +
  theme_minimal()

p
```


## Null vs Dirichlet plots
```{r null vs dirichlet}
# Dirichlet factors

dirichlet_difference_summary <- model_difference_summary %>%
  filter(grepl("^Dirichlet", Model)) %>%
  mutate(
    Factor = case_when(
      Model == "Dirichlet model" ~ 1,
      TRUE ~ as.numeric(
        sub("Dirichlet model ", "", Model)
      )
    ),
    Factor = factor(
      Factor,
      levels = c(1, 2, 3, 4, 5,6,7,8,9, 10, 20, 50)
    )
  )

# Remove the large ones
dirichlet_difference_summary <- dirichlet_difference_summary %>% 
  filter(
    !Model %in% c("Dirichlet model 20", "Dirichlet model 50")
  )


# ---------------------------------------------------------
# Dirichlet plot

p_dirichlet <- ggplot(
  dirichlet_difference_summary,
  aes(
    x = Factor,
    y = Mean_Difference,
    colour = Factor
  )
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.7,
    colour = "grey50"
  ) +
  geom_errorbar(
    aes(
      ymin = CI_lower,
      ymax = CI_upper
    ),
    width = 0.15,
    linewidth = 0.7
  ) +
  geom_point(
    size = 3
  ) +
  facet_wrap(
  ~ factor(
    Quarter,
    levels = c(1, 2, 3),
    labels = c("QT", "HT", "3QT")
  ),
  scales = "free_y"
) +
  scale_colour_viridis_d(option = "C") +
  labs(
    title = "Dirichlet Forecast Error Relative to the Null Model",
    subtitle = "Difference in mean total absolute error with 95% confidence intervals",
    x = "Dirichlet prior concentration factor",
    y = "Change in Mean Total Absolute Error",
    colour = "Prior factor"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14
    ),
    plot.subtitle = element_text(
      size = 10
    ),
    axis.title = element_text(
      face = "bold"
    ),
    axis.text = element_text(
      colour = "black"
    ),
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )

p_dirichlet

```


## Null vs EWMA plots
```{r null vs ewma}
# EWMA factors

ewma_difference_summary <- model_difference_summary %>%
  filter(grepl("^EWMA", Model)) %>%
  mutate(
    Factor = case_when(
      Model == "EWMA model 0.001"   ~ 0.001,
      Model == "EWMA model 0.002"   ~ 0.002,
      Model == "EWMA model 0.003"   ~ 0.003,
      Model == "EWMA model 0.004"   ~ 0.004,
      Model == "EWMA model 0.005"   ~ 0.005,
      Model == "EWMA model 0.006"   ~ 0.006,
      Model == "EWMA model 0.007"   ~ 0.007,
      Model == "EWMA model 0.008"   ~ 0.008,
      Model == "EWMA model 0.01"  ~ 0.01,
      Model == "EWMA model 0.02" ~ 0.02,
      Model == "EWMA model 0.05" ~ 0.05,
      Model == "EWMA model 0.1" ~ 0.1,
      TRUE ~ NA_real_
    ),
    Factor = factor(
      Factor,
      levels = c(0.001, 0.002,0.003,0.004, 0.005,0.006,0.007,0.008, 0.01, 0.02, 0.05, 0.1)
    )
  )
ewma_difference_summary

# Remove the large ones
ewma_difference_summary <- ewma_difference_summary %>% 
  filter(
    !Model %in% c("EWMA model 0.1", "EWMA model 0.05")
  )


# ---------------------------------------------------------
# EWMA plot

p_ewma <- ggplot(
  ewma_difference_summary,
  aes(
    x = Factor,
    y = Mean_Difference,
    colour = Factor
  )
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.7,
    colour = "grey50"
  ) +
  geom_errorbar(
    aes(
      ymin = CI_lower,
      ymax = CI_upper
    ),
    width = 0.15,
    linewidth = 0.7
  ) +
  geom_point(
    size = 2
  ) +
  facet_wrap(
  ~ factor(
    Quarter,
    levels = c(1, 2, 3),
    labels = c("QT", "HT", "3QT")
  ),
  scales = "free_y"
) +
  scale_colour_viridis_d(option = "C") +
  labs(
    title = "EWMA Forecast Error Relative to the Null Model",
    subtitle = "Difference in mean total absolute error with 95% confidence intervals",
    x = expression(EWMA~factor~beta),
    y = "Change in Mean Total Absolute Error",
    colour = expression(beta)
  ) +
  theme_minimal(base_size = 11) +
  theme_minimal(base_size = 11) +
theme(
  plot.title = element_text(
    face = "bold",
    size = 14
  ),
  plot.subtitle = element_text(
    size = 10
  ),
  axis.title = element_text(
    face = "bold"
  ),
  axis.text = element_text(
    colour = "black"
  ),
  axis.text.x = element_text(
    angle = 45,
    hjust = 1
  ),
  strip.text = element_text(
    face = "bold",
    size = 11
  ),
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank(),
  legend.position = "bottom"
)

p_ewma

```

## Null vs QbyQ plots

```{r null vs qbyq}
# QbyQ factors

qbyq_difference_summary <- model_difference_summary %>%
  filter(grepl("^QbyQ", Model)) %>%
  mutate(
    Factor = case_when(
      Model == "QbyQ model 0"     ~ 0,
      Model == "QbyQ model 0.001" ~ 0.001,
      Model == "QbyQ model 0.002" ~ 0.002,
      Model == "QbyQ model 0.003" ~ 0.003,
      Model == "QbyQ model 0.004" ~ 0.004,
      Model == "QbyQ model 0.005" ~ 0.005,
      Model == "QbyQ model 0.006" ~ 0.006,
      Model == "QbyQ model 0.007" ~ 0.007,
      Model == "QbyQ model 0.008" ~ 0.008,
      Model == "QbyQ model 0.01"  ~ 0.01,
      Model == "QbyQ model 0.02"  ~ 0.02,
      TRUE ~ NA_real_
    ),
    Factor = factor(
      Factor,
      levels = c(0, 0.001, 0.002,0.003,0.004, 0.005,0.006,0.007,0.008, 0.01, 0.02)
    )
  )

qbyq_difference_summary


# ---------------------------------------------------------
# QbyQ plot

p_qbyq <- ggplot(
  qbyq_difference_summary,
  aes(
    x = Factor,
    y = Mean_Difference,
    colour = Factor
  )
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.7,
    colour = "grey50"
  ) +
  geom_errorbar(
    aes(
      ymin = CI_lower,
      ymax = CI_upper
    ),
    width = 0.15,
    linewidth = 0.7
  ) +
  geom_point(
    size = 2
  ) +
  facet_wrap(
  ~ factor(
    Quarter,
    levels = c(1, 2, 3),
    labels = c("QT", "HT", "3QT")
  ),
  scales = "free_y"
) +
  scale_colour_viridis_d(option = "C") +
  labs(
    title = "QbyQ Forecast Error Relative to the Null Model",
    subtitle = "Difference in mean total absolute error with 95% confidence intervals",
    x = expression(QbyQ~factor~beta),
    y = "Change in Mean Total Absolute Error",
    colour = expression(beta)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14
    ),
    plot.subtitle = element_text(
      size = 10
    ),
    axis.title = element_text(
      face = "bold"
    ),
    axis.text = element_text(
      colour = "black"
    ),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )

p_qbyq

```

# Section 7.5
## Best models vs null, plots pvals, prop inproved and tables

```{r best models vs null pvals}

overall_summary <- overall_summary %>%
  mutate(
    Model_Type = case_when(
      grepl("^Dirichlet", Model) ~ "Dirichlet",
      grepl("^EWMA", Model) ~ "EWMA",
      grepl("^Extrapolation", Model) ~ "Extrapolation",
      grepl("^Null", Model) ~ "Null",
      grepl("^QbyQ", Model) ~ "QbyQ",
      TRUE ~ Model
    ))

# Select the best model within each model type
best_models_overall <- overall_summary %>%
  group_by(Model_Type) %>%
  slice_min(Overall_MAE, n = 1, with_ties = FALSE) %>%
  ungroup()

(best_models_quarters <- error_summary[which(error_summary$Model %in% best_models_overall$Model),])

best_models_overall

# Calculate paired differences relative to the null model
model_differences <- comparison_errors %>%
  select(Match_id, Quarter, Model, Total_Abs_Error) %>%
  left_join(
    comparison_errors %>%
      filter(Model == "Null model") %>%
      select(Match_id, Quarter, Null_Error = Total_Abs_Error),
    by = c("Match_id", "Quarter")
  ) %>%
   filter(Model %in% best_models_overall$Model) %>%
  mutate(
    Difference = Total_Abs_Error - Null_Error
  )

# One-sided paired t-tests
significance_summary <- model_differences %>%
  group_by(Model, Quarter) %>%
  summarise(
    Mean_Difference = mean(Difference, na.rm = TRUE),
    SD = sd(Difference, na.rm = TRUE),
    n = sum(!is.na(Difference)),
    SE = SD / sqrt(n),
    t_statistic = Mean_Difference / SE,
    p_value = pt(t_statistic, df = n - 1),
    CI_lower = Mean_Difference -
      qt(0.975, df = n - 1) * SE,
    CI_upper = Mean_Difference +
      qt(0.975, df = n - 1) * SE,
    .groups = "drop"
  ) %>%
  mutate(
    p_adjusted = p.adjust(p_value, method = "holm"),
    Significant = p_adjusted < 0.05
  )

significance_summary

# Keep only the selected models
plot_data <- model_differences %>%
  filter(Model %in% best_models_overall$Model) %>%
  filter(!(Model %in% "Null model")) %>%
  filter(!(Model %in% "Extrapolation model"))


# ---------------------------------------------------------
# Distribution of difference in absolute error

p <- ggplot(plot_data, aes(x = Difference, fill = Model)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(
    Model ~ Quarter,
    scales = "free",
    labeller = labeller(
      Quarter = c(
        `1` = "QT",
        `2` = "HT",
        `3` = "3QT"
      )
    )
  ) +
  labs(
    title = "Distribution of Difference in Absolute Error Relative to the Null Model",
    x = "Difference in absolute error (model - null)",
    y = "Number of matches"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
p

# ---------------------------------------------------------
# Proportion of matches where model outperforms null

proportion_improved <- plot_data %>%
  group_by(Model, Quarter) %>%
  summarise(
    Proportion_Improved = mean(Difference < 0, na.rm = TRUE),
    Percentage_Improved = 100 * Proportion_Improved,
    n = sum(!is.na(Difference)),
    .groups = "drop"
  )

proportion_improved

#___________________________________________________________________________
# Combine home and away forecasting errors
error_data <- comparison_errors %>%
  filter(Model %in% best_models_overall$Model) %>%
  filter(!(Model %in% "Extrapolation model")) %>%
  select(Match_id, Quarter, Model, Home_Error, Away_Error) %>%
  tidyr::pivot_longer(
    cols = c(Home_Error, Away_Error),
    names_to = "Team",
    values_to = "Error"
  )

error_summary2 <- error_data %>%
  group_by(Model, Quarter) %>%
  summarise(
    Mean = mean(Error, na.rm = TRUE),
    SD = sd(Error, na.rm = TRUE),
    Q1 = quantile(Error, 0.25, na.rm = TRUE),
    Median = median(Error, na.rm = TRUE),
    Q3 = quantile(Error, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Quarter = recode(
      Quarter,
      `1` = "QT",
      `2` = "HT",
      `3` = "3QT"
    )
  )

error_summary2

```


# Section 7.6
## Bias of best models

```{r best models bias}

# add qbyq 0 for bias comparison
best_models_quarters <- bind_rows(
  best_models_quarters,
  error_summary %>%
    filter(Model == "QbyQ model 0")
)

best_models_quarters

best_models_overall <- bind_rows(
  best_models_overall,
  overall_summary %>%
    filter(Model == "QbyQ model 0")
)

best_models_overall

```


# Section 7.7

## Tables of relative improvement
```{r table of relative improvement}
# Mean MAE across home and away teams
best_models_mae <- best_models_quarters %>%
  mutate(
    Mean_MAE = (Home_MAE + Away_MAE) / 2
  )

# relative difference from null model
relative_null_mae <- best_models_mae %>%
  select(Model, Quarter, Mean_MAE) %>%
  pivot_wider(
    names_from = Quarter,
    values_from = Mean_MAE
  ) %>%
  left_join(
    best_models_mae %>%
      filter(Model == "Null model") %>%
      select(Quarter, Null_MAE = Mean_MAE) %>%
      pivot_wider(
        names_from = Quarter,
        values_from = Null_MAE,
        names_prefix = "Null_"
      ),
    by = character()
  ) %>%
  mutate(
    QT = (`1` - Null_1) / Null_1 * 100,
    HT = (`2` - Null_2) / Null_2 * 100,
    `3QT` = (`3` - Null_3) / Null_3 * 100
  ) %>%
  select(Model, QT, HT, `3QT`) %>%
  mutate(
    across(c(QT, HT, `3QT`), ~ round(.x, 2))
  )

relative_null_mae

# relative improvement between quarter breaks
mae_improvement <- best_models_mae %>%
  select(Model, Quarter, Mean_MAE) %>%
  pivot_wider(
    names_from = Quarter,
    values_from = Mean_MAE
  ) %>%
  mutate(
    `QT -> HT` = (`1` - `2`) / `1` * 100,
    `HT -> 3QT` = (`2` - `3`) / `2` * 100,
    `QT -> 3QT` = (`1` - `3`) / `1` * 100
  ) %>%
  select(Model, `QT -> HT`, `HT -> 3QT`, `QT -> 3QT`) %>%
  mutate(
    across(-Model, ~ round(.x, 2))
  )

mae_improvement


```

# Section 7.9

## nreps

```{r nreps}
#n_values <- c(10,25,100,250,500,1000)
n_values <- c(2000, 5000, 10000)

test_nreps <- function(n_values, chain_starts, alpha, scaled){

  results <- data.frame(
  nreps = integer(),
  Quarter = integer(),
  runtime = numeric(),
  MAE = numeric(),
  SD_Error = numeric()
  )

  for (n in n_values){

    cat("Running n =", n, "\n")

    start <- Sys.time()

    forecasts <- suppressWarnings(season_simulator(chain_starts,
                                                   n,
                                                   alpha,
                                                   scaled))
    end <- Sys.time()

    runtime <- as.numeric(
      difftime(end, start, units = "mins")
    )

    error_summary <- forecasts %>%
      filter(Quarter < 4) %>%
      group_by(Quarter) %>%
      summarise(
        MAE = mean(Score_MAE),
        SD_Error = sd(Score_MAE)
      )
    
    
    results <- rbind(
      results,
      data.frame(
        nreps = n,
        Quarter = error_summary$Quarter,
        runtime = runtime,
        MAE = error_summary$MAE,
        SD_Error = error_summary$SD_Error
      )
)
  }

  return(results)
}

(nreps_results <- test_nreps(n_values, chain_starts2, 0, FALSE))
```


# Section 7.11

## Do holding times vary throughout a match
```{r holing times qbyq}
holding_time_tests <- chain_starts2 %>%
  group_by(pos_and_loc) %>%
  filter(pos_and_loc %in% c(
    "A_A50","A_M","A_H50",
    "H_A50","H_M","H_H50"
  )) %>%
  summarise(
    p_value = kruskal.test(pos_and_loc_time ~ factor(Quarter))$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adjusted = p.adjust(p_value, method = "holm")
  )

holding_time_tests
```

